---
title: "ANTH418 Week 4 Practice Workbook"
subtitle: "Measures of Central Tendency and Dispersion"
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-tip title="Start here — download the editable workbook"}
This webpage is the rendered version of an **editable Quarto workbook**.

1. Download the **Week 4 Practice Package**.
2. Unzip it.
3. Open `week04-practice.qmd` in RStudio.
4. Keep the `data` folder beside the QMD.
5. Work from top to bottom.
6. For each exercise: **predict → run → inspect → explain**.
7. Edit code and write explanations directly in the QMD.
8. Render frequently.
9. Finish with the **Clean Render Test**.

**Do not copy this webpage into RStudio.** The downloadable QMD is the editable source.

[**Download Week 4 Practice Package (.zip)**](../downloads/practice-workbooks/ANTH418_Week04_Practice.zip)
:::

## How to use this workbook

Week 3 established that measurement and research design determine what numerical summaries can legitimately mean. Week 4 asks:

> **Where is the center of a distribution, how widely do observations vary, and which summaries describe the empirical pattern responsibly?**

> **inspect the distribution → locate the center → quantify spread → check resistance → compare summaries → interpret the empirical pattern**

The governing principle is **describe before explaining**.

## Learning objectives

By the end, you should be able to calculate and interpret mode, median, mean, trimmed mean, range, IQR, variance, standard deviation, MAD, and CV; inspect distributions with histograms and boxplots; choose defensible summaries; calculate grouped descriptives; and separate statistical description from archaeological explanation.

# Part I — Measurement level comes first

## 1. Match the statistic to the variable

| Measurement level | Useful center | Useful spread |
|---|---|---|
| Nominal | Mode | counts / proportions |
| Ordinal | Median, mode | IQR, range |
| Interval | Mean, median, mode | SD, variance, IQR |
| Ratio | Mean, median, mode | SD, variance, IQR, CV |

### Practice 1

For artifact raw material, preservation rank, Celsius temperature, projectile-point length, grave-good count, and human stable-isotope measurement, identify the measurement level and a defensible center/spread summary. Explain why software being able to calculate a statistic does not make it meaningful.

# Part II — Central tendency

## 2. Mode

```{r}
library(tidyverse)

artifact_material <- c("chert", "chert", "quartz", "chert",
                       "quartzite", "quartz", "chert")
table(artifact_material)
```

What is the mode? Why is mode useful for nominal variables?

## 3. Median and resistance

```{r}
x <- c(4, 5, 5, 6, 7, 8, 9)
x_extreme <- c(4, 5, 5, 6, 7, 8, 50)

median(x)
median(x_extreme)
```

Predict before running: how much will the median change?

## 4. Mean and sensitivity

```{r}
mean(x)
mean(x_extreme)
```

Compare the mean's change with the median's change. Explain **resistance** in your own words.

## 5. Mean as balance point

```{r}
xbar <- mean(x)
x - xbar
sum(x - xbar)
```

Why do deviations from the arithmetic mean sum to approximately zero?

## 6. Trimmed mean

```{r}
skewed_values <- c(10, 11, 11, 12, 12, 13, 13, 14, 15, 100)

mean(skewed_values)
median(skewed_values)
mean(skewed_values, trim = 0.10)
```

Place the ordinary mean, trimmed mean, and median in order and explain what the comparison reveals.

# Part III — Center alone is incomplete

## 7. Same center, different spread

```{r}
low_spread <- c(8, 9, 10, 10, 10, 11, 12)
wide_spread <- c(1, 5, 8, 10, 12, 15, 19)

mean(low_spread); mean(wide_spread)
median(low_spread); median(wide_spread)
sd(low_spread); sd(wide_spread)
```

Explain why reporting only a mean can hide important structure.

# Part IV — Range and IQR

## 8. Range is fragile

```{r}
range_demo <- c(4, 5, 6, 7, 8)
range_demo_extreme <- c(4, 5, 6, 7, 80)

diff(range(range_demo))
diff(range(range_demo_extreme))
```

Why does range depend so strongly on extremes?

## 9. IQR describes the middle 50%

```{r}
IQR(range_demo)
IQR(range_demo_extreme)
quantile(range_demo_extreme, probs = c(.25, .50, .75))
```

Why does IQR pair naturally with the median?

# Part V — Variance and standard deviation

## 10. Deviations and variance

```{r}
values <- c(6, 8, 10, 12, 14)
xbar <- mean(values)

deviation_table <- tibble(
  value = values,
  deviation = value - xbar,
  squared_deviation = (value - xbar)^2
)

deviation_table
sum(deviation_table$deviation)
sum(deviation_table$squared_deviation)
var(values)
```

Why are deviations squared?

## 11. Standard deviation returns to original units

```{r}
var(values)
sd(values)
sqrt(var(values))
```

Explain the units of variance versus standard deviation.

# Part VI — MAD and robust spread

## 12. Median absolute deviation

```{r}
robust_values <- c(10, 11, 11, 12, 12, 13, 13, 14, 100)
med <- median(robust_values)

median(abs(robust_values - med))
mad(robust_values, constant = 1)
sd(robust_values)
IQR(robust_values)
```

Which spread measures are most affected by the extreme value?

# Part VII — Coefficient of variation

## 13. Relative variability

\[
CV = \frac{s}{\bar{x}}\times 100\%.
\]

```{r}
group_A <- c(9, 10, 11)
group_B <- c(90, 100, 110)

sd(group_A); sd(group_B)
100 * sd(group_A) / mean(group_A)
100 * sd(group_B) / mean(group_B)
```

What does CV add beyond SD?

::: {.callout-important}
Use CV only when its scale and denominator are substantively defensible. Do not calculate it mechanically.
:::

# Part VIII — Distribution shape guides summary choice

## 14. Symmetry and skew

```{r}
symmetric <- c(6, 7, 8, 9, 10, 11, 12)
right_skewed <- c(6, 7, 8, 9, 10, 11, 30)

tibble(
  distribution = c("symmetric", "right_skewed"),
  mean = c(mean(symmetric), mean(right_skewed)),
  median = c(median(symmetric), median(right_skewed))
)
```

```{r}
shape_demo <- tibble(
  value = c(symmetric, right_skewed),
  distribution = rep(c("symmetric", "right_skewed"), each = 7)
)

ggplot(shape_demo, aes(x = value)) +
  geom_histogram(binwidth = 2, boundary = 0) +
  facet_wrap(~ distribution, scales = "free_y") +
  theme_minimal()
```

For symmetric versus strongly skewed distributions, which center-and-spread pairings would you consider, and why?

# Part IX — Boxplots and extreme observations

## 15. Center and spread together

```{r}
ggplot(shape_demo, aes(x = distribution, y = value)) +
  geom_boxplot() +
  theme_minimal()
```

Why should an observation flagged by a boxplot **not** automatically be deleted? List at least three possibilities it could represent.

# Part X — Load the Knipper teaching data

## 16. Portable data path

```{r}
if (file.exists("data/knipper2020_humans.csv")) {
  knipper_file <- "data/knipper2020_humans.csv"
} else {
  project_root <- if (file.exists("_quarto.yml")) "." else ".."
  knipper_file <- file.path(
    project_root, "assignments", "assignment01_knipper_descriptive",
    "data", "knipper2020_humans.csv"
  )
}

file.exists(knipper_file)
```

```{r}
knipper <- read_csv(knipper_file, show_col_types = FALSE)

dim(knipper)
names(knipper)
glimpse(knipper)
head(knipper)
colSums(is.na(knipper))
```

Before summarizing, identify the unit of analysis, sites, isotope variables, and missingness.

# Part XI — Inspect distributions before summarizing

## 17. Reshape the isotope variables

```{r}
iso_long <- knipper |>
  select(site, d13c, d15n) |>
  pivot_longer(
    cols = c(d13c, d15n),
    names_to = "isotope",
    values_to = "value"
  )

glimpse(iso_long)
```

## 18. Histograms

```{r}
ggplot(iso_long, aes(x = value)) +
  geom_histogram(bins = 12) +
  facet_grid(isotope ~ site, scales = "free_x") +
  theme_minimal()
```

Describe symmetry/skew, clustering, possible extremes, site differences, and isotope differences **before** reading summary numbers.

# Part XII — Grouped descriptive statistics

## 19. Center and spread together

```{r}
summary_tbl <- iso_long |>
  group_by(site, isotope) |>
  summarise(
    n = sum(!is.na(value)),
    mean = mean(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    q1 = quantile(value, 0.25, na.rm = TRUE),
    q3 = quantile(value, 0.75, na.rm = TRUE),
    iqr = IQR(value, na.rm = TRUE),
    mad = mad(value, constant = 1, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    range = max(value, na.rm = TRUE) - min(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(
    c(mean, median, sd, q1, q3, iqr, mad, min, max, range),
    ~ round(.x, 2)
  ))

summary_tbl
```

### Practice

For each site × isotope combination, compare mean/median and SD/IQR; inspect min/max; choose a defensible center-and-spread pairing; justify it using numbers **and** distribution shape.

# Part XIII — Boxplots of the case study

## 20. Compare sites visually

```{r}
ggplot(iso_long, aes(x = site, y = value)) +
  geom_boxplot() +
  facet_wrap(~ isotope, scales = "free_y") +
  theme_minimal()
```

Specify which isotope, which center, which spread, approximate magnitude, and whether extreme observations influence the comparison.

# Part XIV — Inspect extreme observations

## 21. Do not delete first and ask later

```{r}
knipper |>
  arrange(d15n) |>
  select(site, d13c, d15n) |>
  slice(c(1:3, (n()-2):n()))
```

For an extreme isotope observation, ask whether it is a data-entry error, measurement problem, genuine biological observation, archaeologically meaningful value, or influential point. Explain why **outlier is a statistical description, not an archaeological explanation**.

# Part XV — Description versus archaeological interpretation

## 22. Keep two tasks separate

Write two paragraphs:

**A — Statistical description:** Describe human isotope distributions at La Bastida and Gatas using center, spread, sample size, and visual shape.

**B — Archaeological interpretation:** State what contextual evidence would be required before explaining the pattern in terms of diet, farming, manuring, or social organization.

# Part XVI — Integrated descriptive analysis

## 23. Mini-analysis challenge

Using the Knipper teaching data:

1. Inspect structure and missingness.
2. Reshape `d13c` and `d15n`.
3. Plot distributions.
4. Calculate `n`, mean, median, SD, Q1, Q3, IQR, minimum, maximum, and range.
5. Compare mean with median and SD with IQR.
6. Inspect possible extreme observations.
7. Choose defensible summaries.
8. Write a concise empirical description.
9. State explicitly what the descriptive statistics **do not establish**.
10. Render the document.

# Part XVII — Conceptual debugging

## 24. Diagnose the mistake

Identify the problem:

1. “The mean is always the best center.”
2. “Same mean means same distribution.”
3. “Range is robust.”
4. “The boxplot flagged it, so I deleted it.”
5. “SD = 2.4, therefore variance = 2.4.”
6. “La Bastida's mean is higher, therefore people there necessarily ate more meat.”
7. “Different sample sizes make descriptive comparison impossible.”
8. “R calculated it, so it must be meaningful.”

# Part XVIII — Mastery checklist

- [ ] I can match measurement level to descriptive summaries.
- [ ] I can interpret mode, median, mean, and trimmed mean.
- [ ] I understand resistance.
- [ ] I can interpret range and IQR.
- [ ] I can explain variance and SD.
- [ ] I can calculate unscaled MAD.
- [ ] I understand CV and its limitations.
- [ ] I inspect histograms and boxplots before choosing summaries.
- [ ] I can calculate grouped descriptives with `dplyr`.
- [ ] I compare center and spread together.
- [ ] I do not automatically delete extreme observations.
- [ ] I separate empirical description from archaeological explanation.

# Final challenge — explain the workflow without code

Explain:

> **measurement level → distribution → center → spread → resistance → visualization → descriptive summary → archaeological interpretation**

Why does Week 4 insist on **describe before explaining**?

# Clean Render Test

1. Save the QMD.
2. Choose **Session → Restart R**.
3. Do not recreate objects manually in the Console.
4. Render the complete document.
5. If it fails, identify what the document assumed but never created.

A successful clean render shows that the descriptive analysis is computationally self-contained.
