---
title: "ANTH418 Week 11 Practice Workbook"
subtitle: "Chi-Square and Tests of Proportions"
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-tip title="Start here — download the editable workbook"}
This webpage is the rendered version of an **editable Quarto workbook**.

1. Download the **Week 11 Practice Package**.
2. Unzip it.
3. Open `week11-practice.qmd` in RStudio.
4. Keep the `data` folder beside the QMD.
5. Work from top to bottom.
6. For each exercise: **predict → run → inspect → explain**.
7. Edit code and write explanations directly in the QMD.
8. Render frequently.
9. Finish with the **Clean Render Test**.

**Do not copy this webpage into RStudio.** The downloadable QMD is the editable source.

[**Download Week 11 Practice Package (.zip)**](../downloads/practice-workbooks/ANTH418_Week11_Practice.zip)
:::

# How to use this workbook

Week 9 compared numerical outcomes between groups. Week 11 changes the statistical object: the evidence is **categorical counts**.

> **define categories → identify units → count → proportions → visualize → expected counts → observed vs. expected → residuals → effect size → archaeology**

> **Week 11 principle: A chi-square test is one part of a larger argument. Define categories, compare observed counts with an explicit expectation, then return to archaeology.**

# Learning objectives

By the end, you should be able to distinguish categorical from numerical variables; identify the unit of analysis; distinguish counts from proportions; choose denominators correctly; distinguish goodness-of-fit from independence tests; construct contingency tables; explain expected counts; interpret chi-square; inspect expected counts and standardized residuals; calculate Cramer's V; explain sparse-table alternatives; visualize categorical evidence; and interpret the Lesotho case without confusing association with causation.

# Part I — Categorical evidence begins with units and labels

## 1. Nominal, ordinal, and numerical variables

```{r}
library(tidyverse)

variable_examples <- tibble(
  variable = c("raw material", "cortex coverage", "artifact length"),
  example = c("FineChert / Dolerite / Quartz", "low / medium / high", "42.6 mm"),
  type = c("nominal", "ordinal", "numerical")
)

variable_examples
```

### Practice 1
Why is the mean of codes such as `1 = chert`, `2 = quartz`, `3 = dolerite` meaningless? Why must the observational unit be identified before counting?

# Part II — Counts and proportions answer different questions

## 2. Counts preserve evidential weight

```{r}
counts_demo <- tibble(
  assemblage = c("Layer A", "Layer B"),
  chert = c(120, 48),
  total = c(300, 120)
) |>
  mutate(proportion_chert = chert / total)

counts_demo
```

Both are 40% chert, but they rest on different denominators.

### Practice 2
For every percentage ask: **percentage of what?** Why should counts remain visible?

# Part III — Proportions have sampling uncertainty

## 3. Same proportion, different precision

$$
SE(\hat p)=\sqrt{\frac{\hat p(1-\hat p)}{n}}
$$

```{r}
tibble(n = c(10, 100, 1000), p_hat = .50) |>
  mutate(
    se = sqrt(p_hat * (1 - p_hat) / n),
    lower_approx = p_hat - 1.96 * se,
    upper_approx = p_hat + 1.96 * se
  )
```

### Practice 3
Why is 50% based on 10 observations less precise than 50% based on 1,000?

# Part IV — Differences in proportions are effect magnitudes

## 4. Absolute and relative differences

```{r}
p1 <- .40
p2 <- .25
p1 - p2
(p1 - p2) / p2
```

The first is **15 percentage points**; the second is a **60% relative increase**.

### Practice 4
Why can both statements be correct while communicating different quantities?

# Part V — Contingency tables expose the comparison

## 5. Cross-classify two categorical variables

```{r}
demo_tab <- matrix(
  c(120, 90, 90,
     30, 60, 30),
  nrow = 2, byrow = TRUE,
  dimnames = list(
    Layer = c("A", "B"),
    Material = c("Chert", "Quartz", "Other")
  )
)

demo_tab
addmargins(demo_tab)
prop.table(demo_tab, margin = 1)
```

### Practice 5
Identify a cell count, row total, column total, grand total, and row proportion.

# Part VI — Expected counts come from a null model

## 6. Independence generates expected frequencies

$$
E_{ij}=\frac{(\text{row total}_i)(\text{column total}_j)}{\text{grand total}}
$$

```{r}
demo_chi <- chisq.test(demo_tab)
demo_chi$observed
round(demo_chi$expected, 2)
```

### Practice 6
Verify one expected count manually. Why do expected frequencies not normally mean equal counts?

# Part VII — Chi-square accumulates discrepancies

## 7. Observed versus expected

$$
\chi^2=\sum \frac{(O-E)^2}{E}
$$

```{r}
demo_contrib <- (demo_chi$observed - demo_chi$expected)^2 / demo_chi$expected
round(demo_contrib, 2)
sum(demo_contrib)
unname(demo_chi$statistic)
```

### Practice 7
Which cells contribute most strongly to the demo statistic?

# Part VIII — Goodness-of-fit and independence differ

## 8. One categorical variable versus two

```{r}
observed <- c(A = 18, B = 22, C = 27, D = 33)
chisq.test(observed, p = rep(.25, 4))
```

```{r}
chisq.test(demo_tab)
```

Goodness-of-fit uses $df=k-1$; independence uses $df=(r-1)(c-1)$.

### Practice 8
Explain why the two tests share observed–expected logic but answer different questions.

# Part IX — Assumptions belong to the design

## 9. Before interpreting Pearson's chi-square

Check **counts rather than percentages**, reasonably **independent observations**, adequate **expected counts**, and defensible sampling, recovery, preservation, and classification.

### Practice 9
Why can fragments from the same object or reduction event create pseudo-replication?

# Part X — Residuals locate the pattern

## 10. The overall test does not tell us where

```{r}
round(demo_chi$residuals, 2)
round(demo_chi$stdres, 2)
```

Positive residuals indicate overrepresentation; negative residuals indicate underrepresentation.

### Practice 10
Why inspect residuals after the overall chi-square result?

# Part XI — A p-value is not association strength

## 11. Cramer's V

```{r}
demo_v <- sqrt(
  unname(demo_chi$statistic) /
    (sum(demo_tab) * min(nrow(demo_tab) - 1, ncol(demo_tab) - 1))
)
demo_v
```

### Practice 11
What does the p-value tell you that Cramer's V does not, and vice versa?

# Part XII — Sparse tables may need alternatives

## 12. Fisher and simulation

```{r}
small_tab <- matrix(c(1, 8, 6, 2), nrow = 2, byrow = TRUE)
small_tab
chisq.test(small_tab)$expected
fisher.test(small_tab)
```

```{r}
set.seed(418)
chisq.test(small_tab, simulate.p.value = TRUE, B = 5000)
```

### Practice 12
Why do alternative reference distributions not repair poor sampling, dependence, or bad classification?

# Part XIII — Visualization should show pattern and weight

## 13. Counts and proportions reveal different things

```{r}
demo_long <- as.data.frame(demo_tab) |>
  as_tibble()

names(demo_long) <- c("layer", "material", "n")

ggplot(demo_long, aes(x = material, y = n, fill = layer)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "Count", title = "Counts preserve assemblage size") +
  theme_minimal()
```

```{r}
demo_prop <- demo_long |>
  group_by(layer) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()

ggplot(demo_prop, aes(x = layer, y = proportion, fill = material)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Within-layer proportion",
       title = "Proportions emphasize composition") +
  theme_minimal()
```

### Practice 13
What does the count graph reveal that the proportion graph hides, and vice versa?

# Part XIV — Load the Lesotho teaching data

## 14. Portable data path

```{r}
if (file.exists("data/lesotho_raw_material.csv")) {
  lesotho_file <- "data/lesotho_raw_material.csv"
} else {
  project_root <- if (file.exists("_quarto.yml")) "." else ".."
  lesotho_file <- file.path(
    project_root,
    "assignments",
    "assignment07_lesotho_chisquare",
    "data",
    "lesotho_raw_material.csv"
  )
}

file.exists(lesotho_file)
```

```{r}
lesotho <- read_csv(lesotho_file, show_col_types = FALSE)

dim(lesotho)
names(lesotho)
glimpse(lesotho)
head(lesotho)
```

The teaching population contains **1,060 source observations** across four published archaeological layers: BAS, RBL-CLBRF, RF, and BARF.

::: {.callout-important title="Keep source data and analysis grouping distinct"}
The seven source raw-material categories remain preserved in the CSV. `Other` is **not** a source category; it is created transparently at analysis time.
:::

# Part XV — Audit the categorical evidence

## 15. Reproduce the source structure before recoding

```{r}
lesotho |> count(level)
lesotho |> count(raw_material, sort = TRUE)
lesotho |> count(level, raw_material)
```

Expected layer totals:

> **BAS 234 · RBL-CLBRF 270 · RF 251 · BARF 305**

### Practice 15
Verify that these sum to 1,060. Why inspect the authentic source categories before recoding?

# Part XVI — Create the transparent teaching grouping

## 16. Preserve three categories; combine four as Other

```{r}
lesotho_grouped <- lesotho |>
  mutate(
    material_group = case_when(
      raw_material == "FineChert" ~ "FineChert",
      raw_material == "Dolerite" ~ "Dolerite",
      raw_material == "CrystallineQuartz" ~ "CrystallineQuartz",
      TRUE ~ "Other"
    ),
    level = factor(level, levels = c("BAS", "RBL-CLBRF", "RF", "BARF")),
    material_group = factor(
      material_group,
      levels = c("CrystallineQuartz", "Dolerite", "FineChert", "Other")
    )
  )

lesotho_grouped |> count(level, material_group)
```

`Other` combines Agate, CoarseChert, Quartzite, and Sandstone.

### Practice 16
Why is this transparent collapse preferable to deleting sparse categories? Why should categories not be collapsed merely to obtain significance?

# Part XVII — Build the four-by-four contingency table

## 17. Counts first

```{r}
lesotho_tab <- xtabs(~ level + material_group, data = lesotho_grouped)

lesotho_tab
addmargins(lesotho_tab)
```

Validated table:

```text
            CrystallineQuartz Dolerite FineChert Other
BAS                        12       59       138    25
RBL-CLBRF                  10       36       194    30
RF                         21       56       140    34
BARF                       44       28       201    32
```

### Practice 17
Identify the largest and smallest cell counts. Which layer has the largest denominator?

# Part XVIII — Proportions reveal composition

## 18. Row proportions

```{r}
row_props <- prop.table(lesotho_tab, margin = 1)
round(row_props, 3)
```

```{r}
lesotho_plot <- as.data.frame(lesotho_tab) |>
  as_tibble() |>
  rename(n = Freq) |>
  group_by(level) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()

ggplot(lesotho_plot, aes(x = level, y = proportion, fill = material_group)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Within-layer proportion", fill = "Raw material",
       title = "Raw-material composition across Sehonghong layers") +
  theme_minimal()
```

### Practice 18
Describe the main compositional differences without inferential language.

# Part XIX — Pearson's chi-square test of independence

## 19. Are layer and raw-material composition independent?

```{r}
lesotho_chi <- chisq.test(lesotho_tab)
lesotho_chi
```

Validated teaching result:

> **χ² ≈ 58.20 · df = 9 · p ≈ 2.98 × 10⁻⁹**

### Practice 19
State the null hypothesis in words. Why should the result not be summarized only as “significant”?

# Part XX — Inspect expected counts

## 20. Is the Pearson approximation reasonable?

```{r}
round(lesotho_chi$expected, 2)
min(lesotho_chi$expected)
sum(lesotho_chi$expected < 5)
```

Expected audit:

> **minimum expected count ≈ 19.21 · cells below 5 = 0**

### Practice 20
Why is this check important? Does a small **observed** count automatically invalidate Pearson's test?

# Part XXI — Residuals show where the association lives

## 21. Standardized residuals

```{r}
stdres <- lesotho_chi$stdres
round(stdres, 2)
```

Important validated patterns include:

- BARF × CrystallineQuartz: strongly positive;
- BAS × Dolerite: positive;
- RBL-CLBRF × FineChert: positive;
- BARF × Dolerite: strongly negative;
- RF × FineChert: negative.

### Practice 21
Choose two positive and two negative residuals. Explain each in observed-versus-expected language.

# Part XXII — Visualize observed-versus-expected departures

## 22. Residual heat map

```{r}
residual_df <- as.data.frame(stdres) |>
  as_tibble() |>
  rename(std_residual = Freq)

ggplot(residual_df, aes(x = material_group, y = level, fill = std_residual)) +
  geom_tile() +
  geom_text(aes(label = round(std_residual, 1))) +
  labs(x = NULL, y = NULL, fill = "Std. residual",
       title = "Which cells drive the layer × raw-material association?") +
  theme_minimal()
```

### Practice 22
How does this heat map add information beyond the stacked proportion chart?

# Part XXIII — Quantify association strength

## 23. Cramer's V

```{r}
cramers_v <- sqrt(
  unname(lesotho_chi$statistic) /
    (sum(lesotho_tab) * min(nrow(lesotho_tab) - 1, ncol(lesotho_tab) - 1))
)

cramers_v
```

Validated result:

> **Cramer's V ≈ 0.135**

### Practice 23
Why should a very small p-value and a Cramer's V around 0.135 not be treated as contradictory?

# Part XXIV — Alternatives are conditional, not ritual

## 24. When would Fisher or simulation be preferable?

For this validated four-by-four table, expected counts are comfortably above 5, so Pearson's approximation is not threatened by sparse expected cells.

### Practice 24
Describe a table for which you would consider Fisher's exact test, simulated p-values, or substantively justified category regrouping. Why is “run Fisher whenever any observed count is below 5” the wrong rule?

# Part XXV — Return to Lesotho archaeology

## 25. Association is not archaeological explanation

The statistical result supports an association between **archaeological layer** and **raw-material composition** in the teaching dataset. Residuals show which cells contribute most strongly, and Cramer's V describes the overall strength of association.

It does **not** identify a unique behavioral mechanism.

Possible explanations require archaeological argument involving procurement preference, mobility, source accessibility, exchange, technology, taphonomy, sampling, classification, and other contextual evidence.

The published Lesotho case also draws on independent evidence such as cortex form and technological ratios. The archaeological argument is therefore broader than a p-value.

### Practice 25

Separate your interpretation into four layers:

1. **What was observed** — raw-material counts and proportions across BAS, RBL-CLBRF, RF, and BARF.
2. **Statistical association** — chi-square, expected counts, residuals, and Cramer's V.
3. **Possible archaeological meaning** — procurement, mobility, accessibility, exchange, technology, or formation processes.
4. **What cannot be established** — a unique behavioral or environmental cause from the contingency table alone.

# Part XXVI — Conceptual debugging

## 26. Diagnose the mistake

For each statement, identify the problem.

1. “FineChert was coded 3, so its mean is larger than Dolerite coded 2.”
2. “The layers have similar percentages, so sample sizes do not matter.”
3. “Expected counts should be equal in every cell.”
4. “One observed cell is below 5, so Pearson's chi-square is automatically invalid.”
5. “The p-value is tiny, so the association must be strong.”
6. “The overall chi-square tells us exactly which cells differ.”
7. “Cramer's V tells us which cells drive the association.”
8. “Because layer and raw material are associated, layer caused procurement choice.”
9. “We should collapse categories until the test becomes significant.”
10. “Fisher's exact test repairs dependence among observations.”

# Part XXVII — Integrated practice

## 27. Reconstruct the complete Week 11 argument

Without copying earlier prose, write a short analysis that:

1. identifies the observational unit;
2. identifies the two categorical variables;
3. reports layer totals;
4. explains the analysis-time `Other` grouping;
5. reports the four-by-four contingency table;
6. reports row proportions;
7. states the null hypothesis;
8. reports chi-square and df;
9. reports the expected-count audit;
10. identifies the strongest residual patterns;
11. reports Cramer's V;
12. explains whether sparse-table alternatives are needed;
13. gives an archaeological interpretation;
14. states at least two claims the analysis does not establish.

# Part XXVIII — Mastery checklist

- [ ] I can distinguish nominal, ordinal, and numerical variables.
- [ ] I identify the observational unit before counting.
- [ ] I report counts as well as proportions.
- [ ] I know that every percentage requires a denominator.
- [ ] I distinguish goodness-of-fit from independence tests.
- [ ] I understand how expected counts arise from the null model.
- [ ] I can explain the chi-square statistic conceptually.
- [ ] I inspect expected counts before interpreting Pearson's test.
- [ ] I use standardized residuals to locate departures.
- [ ] I distinguish evidence of association from association strength.
- [ ] I can calculate and interpret Cramer's V.
- [ ] I understand when Fisher or simulation may be useful.
- [ ] I do not collapse categories merely to obtain significance.
- [ ] I distinguish statistical association from archaeological mechanism.
- [ ] I return statistical results to archaeological context.

# Final challenge — explain the workflow without code

Explain:

> **categorical evidence → counts → proportions → contingency table → null model → expected counts → chi-square → residuals → effect size → alternatives → archaeological interpretation**

Why do **categories and observational units come before the test**?

# Clean Render Test

1. Save the QMD.
2. Choose **Session → Restart R**.
3. Do not recreate objects manually in the Console.
4. Render the complete document.
5. If it fails, identify what the document assumed but never created.

A successful clean render shows that the workflow is computationally self-contained.
