---
title: "ANTH418 Week 9 Practice Workbook"
subtitle: "Comparing Two Groups: t-tests"
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-tip title="Start here — download the editable workbook"}
This webpage is the rendered version of an **editable Quarto workbook**.

1. Download the **Week 9 Practice Package**.
2. Unzip it.
3. Open `week09-practice.qmd` in RStudio.
4. Keep the `data` folder beside the QMD.
5. Work from top to bottom.
6. For each exercise: **predict → run → inspect → explain**.
7. Edit code and write explanations directly in the QMD.
8. Render frequently.
9. Finish with the **Clean Render Test**.

**Do not copy this webpage into RStudio.** The downloadable QMD is the editable source.

[**Download Week 9 Practice Package (.zip)**](../downloads/practice-workbooks/ANTH418_Week09_Practice.zip)
:::

## How to use this workbook

Week 8 established the general logic of statistical inference. Week 9 applies that logic to **comparing two groups**.

> **research design → inspect the groups → estimate the difference → estimate SE → calculate t → quantify uncertainty → assess magnitude → diagnose assumptions → choose alternatives if needed → archaeological interpretation**

> **Do not begin with `t.test()`. Begin with the research question, the observational unit, the distributions, and the size and direction of the difference.**

## Learning objectives

By the end, you should be able to distinguish independent and paired designs; calculate and interpret a mean difference; explain the standard error of a difference; interpret the t statistic and degrees of freedom; explain why Welch's t-test is the independent-samples default; inspect assumptions and robustness; interpret a t-based confidence interval; distinguish raw from standardized effect magnitude; calculate Cohen's d and Hedges' g; explain rank-based and permutation alternatives; and apply the workflow to Tell Tweini isotope evidence.

# Part I — Design comes first

## 1. Independent versus paired observations

```{r}
library(tidyverse)

independent_demo <- tibble(
  phase = rep(c("MBA", "LBA"), each = 5),
  delta13c = c(
    -4.8, -5.1, -4.6, -5.3, -4.9,
    -5.7, -5.4, -6.0, -5.6, -5.8
  )
)

paired_demo <- tibble(
  specimen = 1:8,
  observer_1 = c(41.2, 38.7, 44.0, 40.1, 39.4, 42.8, 37.9, 43.2),
  observer_2 = c(41.0, 38.9, 43.7, 40.0, 39.6, 42.5, 38.0, 43.1)
)

independent_demo
paired_demo
```

### Practice 1

Explain why MBA versus LBA plant observations are independent rather than paired. Why are two measurements of the same specimen paired?

# Part II — The effect must have a direction

## 2. Define Group 1 − Group 2

For independent groups:

$$
\widehat{\Delta}=\bar X_1-\bar X_2
$$

For paired observations:

$$
D_i=X_{1i}-X_{2i}, \qquad \bar D=\frac{\sum D_i}{N}
$$

```{r}
mba_mean <- independent_demo |>
  filter(phase == "MBA") |>
  summarise(m = mean(delta13c)) |>
  pull(m)

lba_mean <- independent_demo |>
  filter(phase == "LBA") |>
  summarise(m = mean(delta13c)) |>
  pull(m)

mba_minus_lba <- mba_mean - lba_mean

mba_mean
lba_mean
mba_minus_lba
```

### Practice 2

Interpret the sign of **MBA − LBA**. What changes if the subtraction order is reversed?

# Part III — Standard error belongs to the difference

## 3. Sampling uncertainty for two means

Welch:

$$
SE_{\text{Welch}}=\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}
$$

Paired:

$$
SE_{\bar D}=\frac{s_D}{\sqrt{N}}
$$

```{r}
mba_demo <- independent_demo$delta13c[independent_demo$phase == "MBA"]
lba_demo <- independent_demo$delta13c[independent_demo$phase == "LBA"]

welch_se <- sqrt(
  var(mba_demo) / length(mba_demo) +
    var(lba_demo) / length(lba_demo)
)
welch_se
```

### Practice 3

Why does the standard-error formula depend on the research design?

# Part IV — The t statistic scales signal by uncertainty

## 4. Difference divided by standard error

$$
t=\frac{(\bar X_1-\bar X_2)-0}{SE_{\text{diff}}}
$$

```{r}
t_manual <- mba_minus_lba / welch_se
t_manual
```

### Practice 4

Explain t as **difference relative to uncertainty**. Why can the same raw difference produce different t values?

# Part V — Degrees of freedom shape the reference distribution

## 5. Different designs use different df

Student independent: $df=n_1+n_2-2$

Paired: $df=N_{\text{pairs}}-1$

Welch uses the Welch–Satterthwaite approximation and can return fractional df.

```{r}
welch_demo <- t.test(delta13c ~ phase, data = independent_demo)
welch_demo
```

### Practice 5

Locate t, df, p, CI, and group means in the R output. Why can Welch df be fractional?

# Part VI — Welch is the independent-samples default

## 6. Student pools; Welch separates

```{r}
t.test(delta13c ~ phase, data = independent_demo)
t.test(delta13c ~ phase, data = independent_demo, var.equal = TRUE)
```

Welch does not require equal population variances. Student's independent t-test does.

### Practice 6

Why should Student's test not be chosen simply because an equal-variance pretest failed to reject?

# Part VII — Paired t-tests analyze difference scores

## 7. Pairing changes the information structure

```{r}
paired_demo <- paired_demo |> mutate(diff = observer_1 - observer_2)

paired_demo
mean(paired_demo$diff)
sd(paired_demo$diff) / sqrt(nrow(paired_demo))
```

```{r}
t.test(paired_demo$observer_1, paired_demo$observer_2, paired = TRUE)
```

### Practice 7

Why would treating these measurements as independent throw away information?

# Part VIII — Inspect before testing

## 8. Distributions, outliers, and sample size

```{r}
ggplot(independent_demo, aes(x = phase, y = delta13c)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.08, size = 2) +
  labs(x = NULL, y = "δ13C (‰)") +
  theme_minimal()
```

```{r}
ggplot(independent_demo, aes(sample = delta13c)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ phase, scales = "free") +
  theme_minimal()
```

### Practice 8

With small samples, what can a Q–Q plot tell you? What can it not establish? Why can no normality test diagnose archaeological independence?

# Part IX — Assumptions and robustness

## 9. Four distinct issues

- **Scale:** the response should support meaningful means.
- **Independence:** determined by design and archaeological context.
- **Distribution shape:** severe skew and influential outliers matter especially at small n.
- **Variance structure:** Student assumes equal variances; Welch relaxes this.

::: {.callout-important}
**Independence is not a visual property of a histogram and cannot be rescued by a large sample.** Repeated measurements, fragments from one original object, or observations clustered within one context may contain less independent information than the row count suggests.
:::

### Practice 9

Give an archaeological example in which many rows do not equal many independent observations.

# Part X — Confidence interval for the mean difference

## 10. Estimate first

$$
\text{difference}\pm t_{\text{crit}}\times SE_{\text{diff}}
$$

```{r}
welch_demo$conf.int
```

### Practice 10

Interpret the interval as a range of plausible effect magnitudes under the procedure. Why is “crosses zero / does not cross zero” incomplete?

# Part XI — Raw and standardized effect magnitude

## 11. Original units remain primary

```{r}
n1 <- length(mba_demo)
n2 <- length(lba_demo)
s1 <- sd(mba_demo)
s2 <- sd(lba_demo)

sp <- sqrt(((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2))
cohen_d <- (mean(mba_demo) - mean(lba_demo)) / sp

df_pooled <- n1 + n2 - 2
J <- 1 - 3 / (4 * df_pooled - 1)
hedges_g <- J * cohen_d

cohen_d
hedges_g
```

### Practice 11

Why is the raw isotope difference more archaeologically legible than Cohen's d? What does Hedges' g correct for?

# Part XII — Alternatives depend on question and design

## 12. Rank-based and resampling alternatives

```{r}
wilcox.test(delta13c ~ phase, data = independent_demo, exact = FALSE)
```

```{r}
wilcox.test(paired_demo$observer_1, paired_demo$observer_2, paired = TRUE, exact = FALSE)
```

Permutation tests construct empirical null distributions when exchangeability is defensible.

### Practice 12

Why are Welch t, rank-based tests, and permutation tests not interchangeable rituals?


# Part XIII — Load the Tell Tweini teaching data

## 13. Portable data path

The Week 9 case study uses a focused teaching extract from the published Tell Tweini plant dataset.

```{r}
if (file.exists("data/tell_tweini_ttest.csv")) {
  tweini_file <- "data/tell_tweini_ttest.csv"
} else {
  project_root <- if (file.exists("_quarto.yml")) "." else ".."
  tweini_file <- file.path(project_root, "assignments", "assignment06_tell_tweini_ttest", "data", "tell_tweini_ttest.csv")
}
file.exists(tweini_file)
```

```{r}
tweini <- read_csv(tweini_file, show_col_types = FALSE)
dim(tweini)
names(tweini)
glimpse(tweini)
head(tweini)
```

The teaching dataset has **410 rows and 5 variables**: `source_row_id`, `period`, `taxon`, `measure`, and `isotope_value`.

::: {.callout-important title="Know what is being measured"}
One row represents one plant Δ13C observation in the teaching dataset. The outcome is **Δ13C**, identified by `measure == "delta13c"` and stored numerically in `isotope_value`. It is **not raw measured plant δ13C**.
:::

### Practice 13 — Audit before analysis

Identify the observational unit, period variable, taxon variable, measurement variable, and numerical outcome. Why must the observational unit be understood before running a t-test?

# Part XIV — Verify the archaeological sample structure

## 14. Period, measure, and taxon counts

```{r}
tweini |> count(period)
```

The validated teaching dataset contains **Early Bronze Age 24 · Middle Bronze Age 144 · Late Bronze Age 135 · Iron Age 107**.

```{r}
tweini |> count(measure)
tweini |> count(taxon, sort = TRUE)
```

### Practice 14

Why are these counts part of statistical analysis rather than merely housekeeping? What should you do if they do not reproduce the expected data structure?

# Part XV — Define the Tell Tweini comparison

## 15. Wild ryegrass: Middle Bronze Age versus Late Bronze Age

The research question is:

> **How large is the difference in wild ryegrass (*Lolium cf. multiflorum*) Δ13C between the Middle Bronze Age and Late Bronze Age at Tell Tweini, how uncertain is that difference, and what does a Welch two-sample t-test contribute to its interpretation?**

```{r}
ryegrass <- tweini |>
  filter(
    taxon == "Lolium cf. multiflorum",
    period %in% c("Middle Bronze Age", "Late Bronze Age"),
    measure == "delta13c"
  )
ryegrass |> count(period)
glimpse(ryegrass)
```

You should obtain **Middle Bronze Age: n = 10** and **Late Bronze Age: n = 10**.

Throughout this workbook define the effect as **Middle Bronze Age − Late Bronze Age**. A positive difference therefore means that mean Δ13C is higher in the Middle Bronze Age.

### Practice 15

State the observational unit, numerical outcome, grouping variable, why the groups are independent rather than paired, and the direction of the contrast.

# Part XVI — Visualize and describe before testing

## 16. Every observation matters

```{r}
ggplot(ryegrass, aes(x = period, y = isotope_value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.06, size = 3) +
  labs(x = NULL, y = "Plant Δ13C", title = "Wild ryegrass Δ13C at Tell Tweini") +
  theme_minimal()
```

```{r}
ryegrass_summary <- ryegrass |>
  group_by(period) |>
  summarise(n = n(), mean = mean(isotope_value), sd = sd(isotope_value), median = median(isotope_value), iqr = IQR(isotope_value), .groups = "drop")
ryegrass_summary
```

### Practice 16

Describe center, spread, overlap, unusual observations, and sample size. Do not make an inferential claim from the plot alone.

# Part XVII — Estimate the observed effect

## 17. Middle Bronze Age − Late Bronze Age

```{r}
mba <- ryegrass |> filter(period == "Middle Bronze Age") |> pull(isotope_value)
lba <- ryegrass |> filter(period == "Late Bronze Age") |> pull(isotope_value)
mean_mba <- mean(mba)
mean_lba <- mean(lba)
raw_diff <- mean_mba - mean_lba
mean_mba
mean_lba
raw_diff
```

### Practice 17

Interpret the sign and magnitude before looking at a p-value. Why does the raw Δ13C difference remain archaeologically important even if we later standardize it?

# Part XVIII — Diagnose assumptions without turning them into gates

## 18. Design, distributions, and Shapiro–Wilk diagnostics

```{r}
ggplot(ryegrass, aes(sample = isotope_value)) +
  stat_qq() + stat_qq_line() + facet_wrap(~ period, scales = "free") + theme_minimal()
```

```{r}
shapiro_results <- ryegrass |>
  group_by(period) |>
  summarise(W = unname(shapiro.test(isotope_value)$statistic), p_value = shapiro.test(isotope_value)$p.value, .groups = "drop")
shapiro_results
```

::: {.callout-warning title="Do not turn diagnostics into a pass/fail ritual"}
With **n = 10 per group**, formal normality diagnostics have limited power. A Shapiro–Wilk p-value cannot establish independence and should not mechanically determine whether you are “allowed” to use a t-test.
:::

### Practice 18

Consider independence, distribution shape, unusual observations, group spread, and sample size separately. Explain whether a mean-based comparison is reasonable.

# Part XIX — Welch two-sample t-test

## 19. Primary inferential analysis

To preserve **Middle Bronze Age − Late Bronze Age**, supply the vectors in that order:

```{r}
welch_result <- t.test(mba, lba, var.equal = FALSE)
welch_result
```

```{r}
welch_t <- unname(welch_result$statistic)
welch_df <- unname(welch_result$parameter)
welch_p <- welch_result$p.value
welch_ci <- welch_result$conf.int
welch_t
welch_df
welch_p
welch_ci
```

The t statistic represents the **observed mean difference relative to its estimated standard error**.

### Practice 19

Report both group means, MBA − LBA, the 95% CI, t, Welch–Satterthwaite df, and the two-sided p-value. Do not reduce the result to “significant” or “not significant.”

# Part XX — Effect magnitude

## 20. Raw difference and Cohen's d

```{r}
n_mba <- length(mba)
n_lba <- length(lba)
sd_mba <- sd(mba)
sd_lba <- sd(lba)
pooled_sd <- sqrt(((n_mba - 1) * sd_mba^2 + (n_lba - 1) * sd_lba^2) / (n_mba + n_lba - 2))
cohen_d <- raw_diff / pooled_sd
raw_diff
cohen_d
```

Cohen's d expresses the observed difference in pooled within-group standard-deviation units.

### Optional extension — Hedges' g

```{r}
df_pooled <- n_mba + n_lba - 2
J <- 1 - 3 / (4 * df_pooled - 1)
hedges_g <- J * cohen_d
hedges_g
```

### Practice 20

Why should the raw difference, Cohen's d, and Hedges' g not be treated as interchangeable? Why is the raw Δ13C contrast especially important archaeologically?

# Part XXI — Compare Welch with the pooled Student t-test

## 21. Same observed means, different variance model

```{r}
student_result <- t.test(mba, lba, var.equal = TRUE)
student_result
```

```{r}
comparison <- tibble(
  method = c("Welch", "Student pooled"),
  t = c(unname(welch_result$statistic), unname(student_result$statistic)),
  df = c(unname(welch_result$parameter), unname(student_result$parameter)),
  p_value = c(welch_result$p.value, student_result$p.value),
  ci_lower = c(welch_result$conf.int[1], student_result$conf.int[1]),
  ci_upper = c(welch_result$conf.int[2], student_result$conf.int[2])
)
comparison
```

The **observed mean difference does not change**. What changes is how sampling uncertainty is modeled.

### Practice 21

Compare t, df, p-value, and CI. Does the substantive conclusion materially change? Why is Welch still a useful default when equal population variances cannot safely be assumed?

# Part XXII — Alternatives and archaeological interpretation

## 22. Changing the procedure can change the question

If severe skew, influential extreme observations, or other features made a mean-based t-test difficult to justify, a commonly used rank-based alternative for two independent groups is the **Mann–Whitney/Wilcoxon rank-sum test**. It should not simply be called “the t-test for non-normal data.”

### Practice 22A

Why is “Shapiro–Wilk p < .05 → automatically use Mann–Whitney” too mechanical? What should be considered in addition to a normality diagnostic?

## 23. Return to Tell Tweini archaeology

The measured contrast is a **period difference in wild ryegrass Δ13C**. The isotope pattern may relate to differences in plant water status and growing conditions. Archaeological explanations may involve crop-field association, agricultural water availability, environmental change, or other processes.

> **The t-test does not identify the mechanism that produced the contrast.**

### Practice 22B

Separate your interpretation into four layers:

1. **What was measured** — the observed period difference in wild ryegrass Δ13C.
2. **Statistical inference** — the estimate, CI, Welch t, df, and p-value.
3. **Possible archaeological meaning** — implications for plant water status, growing conditions, agricultural practices, or environmental change.
4. **What cannot be established** — a unique climatic or agricultural cause.

# Part XXIII — Conceptual debugging

## 23. Diagnose the mistake

1. “Two groups means independent samples.”
2. “Equal sample sizes mean the design is paired.”
3. “A large t means the raw effect must be large.”
4. “Degrees of freedom are just a reporting convention.”
5. “Welch is only for data that fail a variance test.”
6. “A nonsignificant variance test proves equal variances.”
7. “If p < .05, the effect is archaeologically important.”
8. “Cohen's d is always more meaningful than the raw isotope difference.”
9. “A valid t-test proves the observations were independent.”
10. “A significant isotope difference establishes its causal mechanism.”

# Part XXIV — Integrated Week 9 challenge

## 24. Build the full two-group analysis

Using the Tell Tweini teaching data:

1. State an archaeological two-group question.
2. Identify the observational unit.
3. Decide whether the design is independent or paired.
4. Define the direction of the contrast.
5. Inspect individual observations and distributions.
6. Report n, mean, SD, median, and IQR.
7. Calculate the raw difference.
8. Explain the appropriate standard error.
9. Run Welch's t-test for an independent comparison.
10. Report t, df, p, and the 95% CI.
11. Calculate a standardized effect and retain the raw effect.
12. Explain whether a rank-based or permutation alternative would answer the same question.
13. Interpret the result archaeologically.
14. State limitations involving independence, recovery, phase, measurement, or causal interpretation.

# Part XXV — Mastery checklist

- [ ] I distinguish independent from paired designs.
- [ ] I define contrast direction before calculating.
- [ ] I understand SE of a mean difference.
- [ ] I interpret t as difference relative to uncertainty.
- [ ] I understand why df shapes the t distribution.
- [ ] I use Welch as the independent-samples default.
- [ ] I know when a paired t-test is appropriate.
- [ ] I inspect distributions rather than relying on one diagnostic test.
- [ ] I interpret a CI as an estimate of effect uncertainty.
- [ ] I report the raw effect in archaeological units.
- [ ] I understand Cohen's d and Hedges' g.
- [ ] I distinguish t-tests from rank-based and permutation alternatives.
- [ ] I do not confuse statistical significance with archaeological importance.
- [ ] I return every statistical result to archaeological context.

# Final challenge — explain the workflow without code

Explain:

> **design → comparison → standard error → t + df → confidence interval → effect magnitude → diagnostics → alternative if needed → archaeological interpretation**

Why does **design come before the test**?

# Clean Render Test

1. Save the QMD.
2. Choose **Session → Restart R**.
3. Do not recreate objects manually in the Console.
4. Render the complete document.
5. If it fails, identify what the document assumed but never created.

A successful clean render shows that the workflow is computationally self-contained.
