---
title: "ANTH418 Week 8 Practice Workbook"
subtitle: "Confidence Intervals and Hypothesis Testing"
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-tip title="Start here — download the editable workbook"}
This webpage is the rendered version of an **editable Quarto workbook**.

1. Download the **Week 8 Practice Package**.
2. Unzip it.
3. Open `week08-practice.qmd` in RStudio.
4. Keep the `data` folder beside the QMD.
5. Work from top to bottom.
6. For each exercise: **predict → run → inspect → explain**.
7. Edit code and write explanations directly in the QMD.
8. Render frequently.
9. Finish with the **Clean Render Test**.

**Do not copy this webpage into RStudio.** The downloadable QMD is the editable source.

[**Download Week 8 Practice Package (.zip)**](../downloads/practice-workbooks/ANTH418_Week08_Practice.zip)
:::

## How to use this workbook

Week 7 showed why statistics vary from sample to sample. Week 8 uses that variability for inference:

> **estimate → quantify uncertainty → specify a null model → construct a null distribution → calculate a p-value → return to effect magnitude and archaeology**

> **Bootstrap:** How much might my estimate vary?

> **Permutation:** What differences would we expect if the null model held?

A bootstrap distribution is a distribution of **estimates**. A permutation distribution is a **null distribution**.

## Learning objectives

By the end, you should be able to distinguish point and interval estimates; construct and interpret bootstrap confidence intervals; distinguish bootstrap from permutation; state null and alternative hypotheses; construct a permutation null distribution; interpret a p-value as a tail probability conditional on a null model; explain what a p-value does not measure; distinguish statistical significance from effect magnitude and archaeological significance; and apply the complete workflow to Teotihuacan leporid δ13C data.

# Part I — Sampling variability creates uncertainty

## 1. From Week 7 to Week 8

```{r}
library(tidyverse)

set.seed(418)
population <- tibble(value = rgamma(10000, shape = 8, scale = 2))
sample_dat <- population |> slice_sample(n = 40)
mean(sample_dat$value)
```

### Practice 1

Why is the sample mean useful but incomplete? What information about uncertainty is missing?

# Part II — Point estimate versus interval estimate

## 2. One number versus a range

```{r}
point_estimate <- mean(sample_dat$value)
point_estimate
```

A point estimate asks **what is the observed estimate?** An interval estimate adds **how uncertain is it under the procedure being used?**

### Practice 2

Why should archaeological interpretation begin with direction and magnitude rather than a significance label?

# Part III — Bootstrap logic

## 3. Resample observations with replacement

```{r}
set.seed(419)

boot_means <- tibble(rep = 1:5000) |>
  mutate(
    boot_mean = map_dbl(
      rep,
      ~ mean(sample(sample_dat$value, nrow(sample_dat), replace = TRUE))
    )
  )

ggplot(boot_means, aes(x = boot_mean)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = point_estimate, linetype = 2) +
  theme_minimal()
```

### Practice 3

What does one bootstrap mean represent? Why is this not a distribution of individual observations?

# Part IV — Bootstrap confidence intervals

## 4. Percentile interval

```{r}
boot_ci <- quantile(boot_means$boot_mean, probs = c(0.025, 0.975))
boot_ci
```

```{r}
ggplot(boot_means, aes(x = boot_mean)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = boot_ci, linetype = "dashed") +
  geom_vline(xintercept = point_estimate) +
  theme_minimal()
```

### Practice 4

Interpret the interval in terms of magnitude and uncertainty. Do **not** say there is a 95% probability that the fixed population parameter lies inside this particular frequentist interval.

# Part V — Confidence level is a property of a procedure

## 5. Repeated-interval thinking

The confidence level refers to long-run performance of the interval-producing procedure under its assumptions.

### Practice 5

Explain why this is preferable to saying “there is a 95% probability that the parameter is inside this particular interval.”

# Part VI — Hypotheses are statistical models

## 6. Null and alternative

For a difference between means:

$$
H_0:\mu_1-\mu_2=0
$$

$$
H_A:\mu_1-\mu_2\neq0
$$

The null is a **statistical reference model**, not necessarily a claim that “nothing happened.”

### Practice 6

Why is “the isotope means differ” not equivalent to “people managed leporids”?

# Part VII — Permutation logic

## 7. Shuffle what the null permits

```{r}
set.seed(420)

perm_demo <- tibble(
  group = rep(c("A", "B"), each = 25),
  value = c(rnorm(25, 10, 2), rnorm(25, 12, 2))
)

obs_diff_demo <- mean(perm_demo$value[perm_demo$group == "B"]) -
  mean(perm_demo$value[perm_demo$group == "A"])

obs_diff_demo
```

```{r}
set.seed(421)

perm_null_demo <- tibble(rep = 1:5000) |>
  mutate(
    stat = map_dbl(
      rep,
      ~ {
        shuffled <- sample(perm_demo$group, replace = FALSE)
        mean(perm_demo$value[shuffled == "B"]) -
          mean(perm_demo$value[shuffled == "A"])
      }
    )
  )

ggplot(perm_null_demo, aes(x = stat)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = c(-abs(obs_diff_demo), abs(obs_diff_demo))) +
  theme_minimal()
```

### Practice 7

Why is the permutation scheme part of the model rather than merely a computational trick?

# Part VIII — The p-value

## 8. Tail probability under the null

```{r}
p_demo <- mean(abs(perm_null_demo$stat) >= abs(obs_diff_demo))
p_demo
```

The p-value asks how unusual a statistic at least this extreme would be **if the null model were an adequate description of the data-generating process**.

### Practice 8

Explain why a p-value is not the probability that H0 is true, the probability that chance caused the result, the effect size, the probability an archaeological explanation is true, or a measure of archaeological importance.

# Part IX — Bootstrap and permutation answer different questions

## 9. Two resampling tools

| Procedure | What is resampled? | Distribution | Week 8 question |
|---|---|---|---|
| Bootstrap | observations with replacement | estimates | How much might my estimate vary? |
| Permutation | labels allowed by H0 | null statistics | What differences would occur if H0 held? |

### Practice 9

Why is only the permutation distribution a null distribution here?

# Part X — Statistical significance is not effect magnitude

## 10. Three separate questions

1. **How large is the observed effect?**
2. **How uncertain is the estimate?**
3. **How unusual is the result under a defined null model?**

A p-value answers only the third.

### Practice 10

Why can a small effect be statistically detectable while an archaeologically important effect can remain uncertain?

# Part XI — Load the Teotihuacan teaching data

## 11. Portable data path

```{r}
if (file.exists("data/teotihuacan_leporids.csv")) {
  data_path <- "data/teotihuacan_leporids.csv"
} else {
  project_root <- if (file.exists("_quarto.yml")) "." else ".."
  data_path <- file.path(
    project_root,
    "assignments",
    "assignment05_teotihuacan_inference",
    "data",
    "teotihuacan_leporids.csv"
  )
}

file.exists(data_path)
```

```{r}
dat <- read_csv(data_path, show_col_types = FALSE) |>
  filter(valid)

glimpse(dat)
names(dat)
```

::: {.callout-important title="Know the archaeological comparison"}
One row is one source observation for a leporid specimen in the teaching file. `context` is the grouping variable and `d13c` is the numerical response in ‰ VPDB. `valid` refers to the δ13C exclusion rule documented for the teaching data.

These specimens are **not automatically a random probability sample of all leporids that lived at Teotihuacan**.
:::

# Part XII — Define the comparison before calculating

## 12. Oztoyahualco versus Moon Pyramid

```{r}
d <- dat |>
  filter(context %in% c("Oztoyahualco", "Moon Pyramid"))

d |> count(context)
```

Throughout, define the effect as:

> **Oztoyahualco − Moon Pyramid**

A positive difference means mean δ13C is higher (less negative) at Oztoyahualco.

### Practice 11

Why must contrast direction be defined before calculation and interpretation?

# Part XIII — Describe before testing

## 13. Group summaries and distributions

```{r}
group_summary <- d |>
  group_by(context) |>
  summarise(
    n = n(),
    mean_d13c = mean(d13c),
    sd_d13c = sd(d13c),
    .groups = "drop"
  )

group_summary
```

```{r}
ggplot(d, aes(x = context, y = d13c)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.65) +
  labs(x = NULL, y = "δ13C (‰ VPDB)") +
  theme_minimal()
```

### Practice 12

Describe direction, overlap, spread, and sample size **before** asking about statistical significance.

# Part XIV — Estimate the observed contrast

## 14. Difference in means

```{r}
mean_ozto <- mean(d$d13c[d$context == "Oztoyahualco"])
mean_moon <- mean(d$d13c[d$context == "Moon Pyramid"])

obs_diff <- mean_ozto - mean_moon

mean_ozto
mean_moon
obs_diff
```

### Practice 13

Interpret the sign and magnitude in ‰. What does the point estimate not tell you about uncertainty?

# Part XV — Bootstrap the Teotihuacan effect

## 15. Resample within contexts

```{r}
set.seed(418)

ozto <- d$d13c[d$context == "Oztoyahualco"]
moon <- d$d13c[d$context == "Moon Pyramid"]

boot_diff <- tibble(rep = 1:10000) |>
  mutate(
    stat = map_dbl(
      rep,
      ~ mean(sample(ozto, length(ozto), replace = TRUE)) -
        mean(sample(moon, length(moon), replace = TRUE))
    )
  )

ci <- quantile(boot_diff$stat, probs = c(0.025, 0.975))

ci
```

```{r}
ggplot(boot_diff, aes(x = stat)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = ci, linetype = "dashed") +
  geom_vline(xintercept = obs_diff) +
  labs(
    x = "Bootstrap difference in means (‰)",
    y = "Count",
    title = "Bootstrap distribution: Oztoyahualco − Moon Pyramid"
  ) +
  theme_minimal()
```

### Practice 14

Report the observed difference and interval. Interpret the interval for **magnitude and uncertainty**, not merely whether it includes zero.

# Part XVI — Specify the Teotihuacan null model

## 16. No systematic contextual difference

$$
H_0:\mu_{\text{Oztoyahualco}}-\mu_{\text{Moon Pyramid}}=0
$$

$$
H_A:\mu_{\text{Oztoyahualco}}-\mu_{\text{Moon Pyramid}}\neq0
$$

Under H0, context carries no systematic information about δ13C for this comparison.

### Practice 15

Why is this not equivalent to saying that no provisioning or management occurred?

# Part XVII — Permutation null distribution

## 17. Shuffle context labels

The observed δ13C values stay fixed; the labels are shuffled while group sizes are preserved.

```{r}
set.seed(418)

n_ozto <- sum(d$context == "Oztoyahualco")

perm_null <- tibble(rep = 1:10000) |>
  mutate(
    stat = map_dbl(
      rep,
      ~ {
        shuffled <- sample(d$d13c, replace = FALSE)
        mean(shuffled[seq_len(n_ozto)]) -
          mean(shuffled[(n_ozto + 1):length(shuffled)])
      }
    )
  )

perm_null
```

```{r}
ggplot(perm_null, aes(x = stat)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = c(-abs(obs_diff), abs(obs_diff))) +
  labs(
    x = "Difference in means under H0 (‰)",
    y = "Count",
    title = "Permutation null distribution"
  ) +
  theme_minimal()
```

# Part XVIII — Permutation p-value

## 18. Two-sided tail probability

```{r}
p_value <- mean(abs(perm_null$stat) >= abs(obs_diff))
p_value
```

### Practice 16

Write one correct sentence interpreting the p-value as a tail probability conditional on H0, and one sentence explaining what it does **not** tell you.

# Part XIX — Integrate estimate, interval, and test

## 19. Three outputs, three roles

```{r}
tibble(
  observed_difference = obs_diff,
  ci_lower = unname(ci[1]),
  ci_upper = unname(ci[2]),
  permutation_p = p_value
)
```

### Practice 17

Answer:

1. How large is the observed contrast?
2. How uncertain is it?
3. What range of differences is represented by the bootstrap interval?
4. How unusual is the observed contrast under H0?
5. Do estimate, interval, and null-model comparison tell a coherent statistical story?

Do not let the p-value replace the effect estimate or uncertainty.

# Part XX — Return to Teotihuacan archaeology

## 20. Statistical contrast is not archaeological explanation

The Week 8 lecture emphasizes that Oztoyahualco leporids have higher mean carbon-isotope values than other sampled contexts. Higher δ13C is consistent with greater C4/CAM consumption. Access to cultivated foods can be consistent with human provisioning.

But:

> **A statistical alternative such as “isotope means differ” is not equivalent to “people managed leporids.”**

Archaeological interpretation requires context, chronology, ecology, alternative explanations, and other evidence.

### Practice 18

Discuss:

- what higher δ13C may imply about diet;
- why cultivated-food access may be consistent with provisioning;
- why provisioning is not automatically management or breeding;
- how archaeological context contributes to interpretation; and
- at least one limitation or alternative explanation.

# Part XXI — Statistical versus archaeological significance

## 21. Return to substantive meaning

Statistical inference evaluates compatibility with a model and uncertainty around estimates. Archaeological significance asks whether the magnitude and pattern matter for understanding past behavior.

### Practice 19

Complete:

> The statistical analysis establishes __________________________.

> The broader archaeological evidence suggests __________________.

Why should these statements not be collapsed into one?

# Part XXII — Conceptual debugging

## 22. Diagnose the mistake

1. “A 95% CI means there is a 95% probability that the parameter is inside this interval.”
2. “The bootstrap distribution is the distribution of specimens.”
3. “Bootstrap and permutation are interchangeable.”
4. “The null hypothesis means nothing happened.”
5. “A p-value is the probability that H0 is true.”
6. “p = .01 means there is a 1% probability that chance caused the result.”
7. “A small p-value means a large effect.”
8. “Statistical significance establishes archaeological importance.”
9. “A significant isotope contrast proves leporid management.”
10. “Once I have a p-value, I no longer need the estimate or confidence interval.”

# Part XXIII — Integrated Week 8 challenge

## 23. Build the full inferential chain

Using the Teotihuacan teaching data:

1. Identify the observational unit.
2. Define the grouping and response variables.
3. State the direction of the contrast.
4. Describe both groups.
5. Calculate the observed difference.
6. Bootstrap the difference.
7. Construct a 95% percentile interval.
8. State H0 and HA.
9. Build the permutation null distribution.
10. Calculate the two-sided p-value.
11. Integrate effect magnitude, uncertainty, and null-model compatibility.
12. Return to archaeological interpretation.
13. State at least one alternative explanation or limitation.
14. Distinguish the statistical result from the archaeological claim.

# Part XXIV — Mastery checklist

- [ ] I distinguish point and interval estimates.
- [ ] I understand what a bootstrap distribution contains.
- [ ] I can construct a percentile bootstrap CI.
- [ ] I can state H0 and HA.
- [ ] I understand permutation as a null-model procedure.
- [ ] I distinguish bootstrap from permutation.
- [ ] I interpret a p-value as a conditional tail probability.
- [ ] I know what a p-value does not measure.
- [ ] I report effect magnitude before significance.
- [ ] I distinguish statistical from archaeological significance.
- [ ] I can integrate estimate, CI, and p-value.
- [ ] I can interpret the Teotihuacan isotope contrast cautiously.

# Final challenge — explain the workflow without code

Explain:

> **estimate → quantify uncertainty → specify H0 → construct a null distribution → calculate a p-value → integrate the evidence → return to archaeology**

Why should **estimation come before significance testing**?

# Clean Render Test

1. Save the QMD.
2. Choose **Session → Restart R**.
3. Do not recreate objects manually in the Console.
4. Render the complete document.
5. If it fails, identify what the document assumed but never created.

A successful clean render shows that the inferential workflow is computationally self-contained.
