---
title: "ANTH418 Week 7 Practice Workbook"
subtitle: "Probability, Sampling, and Sampling Distributions"
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-tip title="Start here — download the editable workbook"}
This webpage is the rendered version of an **editable Quarto workbook**.

1. Download the **Week 7 Practice Package**.
2. Unzip it.
3. Open `week07-practice.qmd` in RStudio.
4. Keep the `data` folder beside the QMD.
5. Work from top to bottom.
6. For each exercise: **predict → run → inspect → explain**.
7. Edit code and write explanations directly in the QMD.
8. Render frequently.
9. Finish with the **Clean Render Test**.

**Do not copy this webpage into RStudio.** The downloadable QMD is the editable source.

[**Download Week 7 Practice Package (.zip)**](../downloads/practice-workbooks/ANTH418_Week07_Practice.zip)
:::

## How to use this workbook

Week 7 asks how partial observations can support defensible inference.

> **population → sample → statistic → repeated sampling → sampling distribution → standard error → archaeological interpretation**

The central distinction is:

> **A data distribution describes observations. A sampling distribution describes how a statistic varies across repeated samples.**

Simulation makes that distinction visible.

## Learning objectives

By the end, you should be able to explain probability as long-run relative frequency; distinguish population, sample, parameter, and statistic; explain major sampling strategies; distinguish sampling error from sampling bias; generate and interpret a sampling distribution; explain standard error and its relation to sample size; explain the Central Limit Theorem at an introductory level; distinguish sample size from sample quality; evaluate selection bias and representativeness; and use p3k14c teaching data to study archaeological sampling.

# Part I — Probability as long-run frequency

## 1. Repeated trials stabilize

```{r}
library(tidyverse)

set.seed(418)

coin_experiment <- tibble(n = c(10, 100, 1000, 10000)) |>
  mutate(
    heads = map_int(n, ~ sum(sample(c(0, 1), .x, replace = TRUE))),
    proportion_heads = heads / n
  )

coin_experiment
```

### Practice 1

Predict which sample size will usually be closest to 0.50. Why does a fair coin not promise exactly five heads in ten flips?

# Part II — Population, sample, parameter, statistic

## 2. Four terms that must not be confused

- **Population:** full target set.
- **Sample:** observed subset.
- **Parameter:** numerical feature of a population.
- **Statistic:** numerical feature calculated from a sample.

```{r}
set.seed(419)

population <- tibble(
  id = 1:10000,
  age = rgamma(10000, shape = 9, scale = 500)
)

population_mean <- mean(population$age)

one_sample <- population |>
  slice_sample(n = 50)

sample_mean <- mean(one_sample$age)

population_mean
sample_mean
```

### Practice 2

Which number is the parameter? Which is the statistic? Why should they differ?

# Part III — Random sampling is a design property

## 3. Simple random sampling

```{r}
set.seed(420)

srs <- population |>
  slice_sample(n = 100)

mean(srs$age)
```

Why is “I chose 100 cases” not enough to establish random sampling? What must you know about selection?

# Part IV — Sampling strategies

## 4. Designs for different constraints

Week 7 distinguishes simple random, stratified, cluster, systematic, convenience, purposive/judgment, snowball, and quota sampling.

### Practice 3

Choose and justify a design for:

1. 1,000 excavation grid units, budget to excavate 50.
2. A regional survey where coastal, valley, and upland zones must all appear.
3. A huge landscape where travel between localities is expensive.
4. An ethnographic study seeking rare specialist knowledge.

Which are probability samples?

# Part V — Sampling error versus sampling bias

## 5. Random samples vary

```{r}
set.seed(421)

five_samples <- tibble(sample_id = 1:5) |>
  mutate(
    sample_mean = map_dbl(
      sample_id,
      ~ population |> slice_sample(n = 50) |> pull(age) |> mean()
    )
  )

five_samples
```

## 6. Bias is different

```{r}
biased_frame <- population |>
  filter(age > median(age))

set.seed(422)

biased_sample <- biased_frame |>
  slice_sample(n = 200)

mean(population$age)
mean(biased_frame$age)
mean(biased_sample$age)
```

### Practice 4

> **Sampling error = random variation around a target.**

> **Sampling bias = systematic displacement caused by selection.**

Why can a large sample from a biased frame remain biased?

# Part VI — Build a sampling distribution

## 7. Repeat the sampling process

```{r}
set.seed(423)

sampling_distribution_50 <- tibble(rep = 1:2000) |>
  mutate(
    sample_mean = map_dbl(
      rep,
      ~ population |> slice_sample(n = 50) |> pull(age) |> mean()
    )
  )

ggplot(sampling_distribution_50, aes(x = sample_mean)) +
  geom_histogram(bins = 35) +
  geom_vline(xintercept = population_mean, linetype = 2) +
  labs(
    x = "Sample mean",
    y = "Number of repeated samples",
    title = "Sampling distribution of the mean, n = 50"
  ) +
  theme_minimal()
```

### Practice 5

What does one value represent? Why is this not a histogram of the 10,000 population observations?

# Part VII — Standard error

## 8. Quantify sample-to-sample variation

```{r}
empirical_se_50 <- sd(sampling_distribution_50$sample_mean)
theoretical_se_50 <- sd(population$age) / sqrt(50)

empirical_se_50
theoretical_se_50
```

Explain the difference among standard deviation, standard error, and one sample mean.

# Part VIII — Sample size changes precision

## 9. Compare n = 20, 50, and 200

```{r}
set.seed(424)

simulate_means <- function(sample_n, reps = 2000) {
  tibble(rep = 1:reps) |>
    mutate(
      n = sample_n,
      sample_mean = map_dbl(
        rep,
        ~ population |>
          slice_sample(n = sample_n) |>
          pull(age) |>
          mean()
      )
    )
}

sampling_sizes <- bind_rows(
  simulate_means(20),
  simulate_means(50),
  simulate_means(200)
)

ggplot(sampling_sizes, aes(x = sample_mean)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ n, scales = "free_y") +
  geom_vline(xintercept = population_mean, linetype = 2) +
  theme_minimal()
```

```{r}
sampling_sizes |>
  group_by(n) |>
  summarise(
    mean_of_means = mean(sample_mean),
    empirical_se = sd(sample_mean),
    .groups = "drop"
  )
```

### Practice 6

What happens to center, spread, and standard error as `n` increases? Why does precision improve approximately with $1/\sqrt{n}$?

# Part IX — Central Limit Theorem

## 10. Sampling distributions can become approximately normal

```{r}
ggplot(population, aes(x = age)) +
  geom_histogram(bins = 40) +
  theme_minimal()
```

```{r}
set.seed(425)

clt_demo <- bind_rows(
  simulate_means(5),
  simulate_means(30),
  simulate_means(100)
)

ggplot(clt_demo, aes(x = sample_mean)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ n, scales = "free") +
  theme_minimal()
```

### Practice 7

State the CLT carefully: it concerns the **sampling distribution of a statistic**, not a claim that raw observations become normal.

# Part X — Sample size is not sample quality

## 11. Large n cannot repair a bad frame

```{r}
set.seed(426)

bias_comparison <- tibble(
  design = c("Random n=50", "Random n=500", "Biased-frame n=500"),
  estimate = c(
    population |> slice_sample(n = 50) |> pull(age) |> mean(),
    population |> slice_sample(n = 500) |> pull(age) |> mean(),
    biased_frame |> slice_sample(n = 500) |> pull(age) |> mean()
  )
)

bias_comparison
```

### Practice 8

Which problem does increasing sample size address most directly: random sampling variability or systematic selection bias?

> **A large sample can be very precisely wrong when the sampling frame is biased.**

# Part XI — Load the p3k14c teaching data

## 12. Portable data path

```{r}
if (file.exists("data/p3k14c_sampling.csv")) {
  p3k_file <- "data/p3k14c_sampling.csv"
} else {
  project_root <- if (file.exists("_quarto.yml")) "." else ".."
  p3k_file <- file.path(
    project_root,
    "assignments",
    "assignment04_p3k14c_sampling",
    "data",
    "p3k14c_sampling.csv"
  )
}

file.exists(p3k_file)
```

```{r}
p3k <- read_csv(p3k_file, show_col_types = FALSE)

dim(p3k)
names(p3k)
glimpse(p3k)
head(p3k)
```

::: {.callout-important title="Know the empirical frame"}
This is the frozen ANTH418 Assignment 4 teaching dataset derived from p3k14c and focused on **Southern California archaeological radiocarbon determinations**. Treat it as the empirical frame for these exercises. It is not all archaeological activity in Southern California and not the entire global p3k14c database.
:::

# Part XII — Audit the radiocarbon frame

## 13. Inspect before sampling

```{r}
summary(p3k)
```

```{r}
p3k |>
  summarise(across(everything(), ~ sum(is.na(.x))))
```

### Practice 9

Using the actual column names, identify the conventional radiocarbon-age variable, geographic/contextual variables, missingness, and the unit represented by one row. State what population this file can and cannot represent.

# Part XIII — One archaeological sample

## 14. Parameter of a frame versus statistic

Inspect `names(p3k)` and assign the actual conventional-age column:

```{r eval=FALSE}
age_var <- "REPLACE_WITH_ACTUAL_AGE_COLUMN"
```

Then:

```{r eval=FALSE}
p3k_complete <- p3k |>
  filter(!is.na(.data[[age_var]]))

frame_mean <- mean(p3k_complete[[age_var]])

set.seed(427)
sample_50 <- p3k_complete |>
  slice_sample(n = 50)

sample_mean_50 <- mean(sample_50[[age_var]])

frame_mean
sample_mean_50
```

### Practice 10

Why is `frame_mean` a parameter **of the supplied teaching frame**, rather than automatically “the true mean age of Southern California archaeology”?

# Part XIV — Sampling distribution with p3k14c

## 15. Repeat random samples

```{r eval=FALSE}
set.seed(428)

p3k_sampling <- tibble(rep = 1:1000) |>
  mutate(
    mean_age = map_dbl(
      rep,
      ~ p3k_complete |>
        slice_sample(n = 50) |>
        pull(all_of(age_var)) |>
        mean()
    )
  )

ggplot(p3k_sampling, aes(x = mean_age)) +
  geom_histogram(bins = 30) +
  geom_vline(
    xintercept = mean(p3k_complete[[age_var]]),
    linetype = 2
  ) +
  theme_minimal()
```

### Practice 11

What does the width of this sampling distribution mean archaeologically?

# Part XV — Sample size and precision in p3k14c

## 16. Compare several sample sizes

```{r eval=FALSE}
simulate_p3k_means <- function(sample_n, reps = 1000) {
  tibble(rep = 1:reps) |>
    mutate(
      n = sample_n,
      mean_age = map_dbl(
        rep,
        ~ p3k_complete |>
          slice_sample(n = sample_n) |>
          pull(all_of(age_var)) |>
          mean()
      )
    )
}

p3k_sizes <- bind_rows(
  simulate_p3k_means(25),
  simulate_p3k_means(50),
  simulate_p3k_means(100)
)

p3k_sizes |>
  group_by(n) |>
  summarise(empirical_se = sd(mean_age), .groups = "drop")
```

### Practice 12

Why does this demonstrate precision **within the available frame**, not correction of archaeological sampling bias?

# Part XVI — Construct a biased archaeological sample

## 17. Selection can move an estimate

Choose an actual geographic or contextual variable from the teaching file and create a deliberately restricted subset.

```{r eval=FALSE}
biased_subset <- p3k_complete |>
  filter(REPLACE_WITH_A_REAL_RESTRICTION)

mean(p3k_complete[[age_var]])
mean(biased_subset[[age_var]])
```

### Practice 13

Is the difference ordinary sampling variability, deliberate selection bias, or both? Does increasing the biased sample size remove the bias?

# Part XVII — p3k14c is large, but not neutral

## 18. Database size is not representativeness

The Week 7 lecture emphasizes that the global p3k14c database is geographically uneven; Europe and North America dominate the compiled radiocarbon determinations.

> **A big database can still be geographically and institutionally uneven.**

### Practice 14

Explain how preservation, excavation history, research funding, laboratory access, publication, database inclusion, and geographic research intensity can shape the observed database before statistical sampling begins.

# Part XVIII — Dates-as-data

## 19. Radiocarbon dates are observations produced by research

A radiocarbon determination is not simply a person or event count.

### Practice 15

Why does “more radiocarbon dates in a period” not automatically mean “more people lived in that period”? What assumptions connect dates-as-data to demographic interpretation?

# Part XIX — Sampling distributions versus chronological distributions

## 20. Two different distributions

A chronological distribution of radiocarbon ages describes **observed dates**.

A sampling distribution describes **how a statistic varies across repeated samples**.

### Practice 16

Give an example of each and explain why they must not be confused.

# Part XX — Simulation as a statistical laboratory

## 21. Why simulation matters

Simulation lets us specify a population or empirical frame, repeat a sampling process, calculate a statistic, and observe long-run behavior.

### Practice 17

Explain why simulation is useful for learning sampling variability, standard error, bias, sample-size effects, and the Central Limit Theorem. What can repeated simulation demonstrate that one observed sample cannot?

# Part XXI — Integrated Week 7 challenge

## 22. Build a sampling argument

Using the p3k14c teaching data:

1. Define the empirical sampling frame.
2. Identify the unit of observation.
3. Identify the conventional radiocarbon-age variable.
4. Calculate the frame mean.
5. Draw one random sample and calculate its mean.
6. Repeat sampling many times.
7. Plot the sampling distribution.
8. Estimate its empirical standard error.
9. Repeat with at least two sample sizes.
10. Construct one deliberately biased subset.
11. Compare precision and bias.
12. Explain why large `n` does not guarantee representativeness.
13. Identify archaeological filters shaping the database.
14. Write an interpretation that distinguishes the empirical frame from the target population.

# Part XXII — Conceptual debugging

## 23. Diagnose the mistake

1. “A random sample should exactly equal the population mean.”
2. “A sample of 1,000 is automatically representative.”
3. “Sampling error and sampling bias are the same.”
4. “A sampling distribution is the histogram of raw observations.”
5. “Standard error measures variation among individual observations.”
6. “The CLT means the raw population becomes normal.”
7. “Increasing `n` removes selection bias.”
8. “The p3k14c teaching file is a random sample of all past human activity.”
9. “More radiocarbon dates necessarily mean more people.”
10. “Simulation is artificial, so it cannot teach us about sampling.”

# Part XXIII — Mastery checklist

- [ ] I understand probability as long-run relative frequency.
- [ ] I distinguish population, sample, parameter, and statistic.
- [ ] I understand that random sampling is a design property.
- [ ] I distinguish probability and nonprobability sampling.
- [ ] I distinguish sampling error from sampling bias.
- [ ] I can generate a sampling distribution.
- [ ] I distinguish a data distribution from a sampling distribution.
- [ ] I understand standard error.
- [ ] I understand how sample size affects precision.
- [ ] I can explain the CLT without saying raw data become normal.
- [ ] I understand that large `n` does not guarantee representativeness.
- [ ] I can audit the p3k14c teaching frame.
- [ ] I understand dates-as-data.
- [ ] I can explain archaeological filters on radiocarbon databases.

# Final challenge — explain the workflow without code

Explain:

> **define the target → identify the frame → sample → calculate a statistic → repeat → observe variability → quantify standard error → audit bias → interpret cautiously**

Why does **representativeness matter more than raw sample size**?

# Clean Render Test

1. Save the QMD.
2. Choose **Session → Restart R**.
3. Do not recreate objects manually in the Console.
4. Render the complete document.
5. If it fails, identify what the document assumed but never created.

A successful clean render shows that the sampling workflow is computationally self-contained.
