---
title: "ANTH418 Week 6 Practice Workbook"
subtitle: "Data Transformation and Missing Values"
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-tip title="Start here — download the editable workbook"}
This webpage is the rendered version of an **editable Quarto workbook**.

1. Download the **Week 6 Practice Package**.
2. Unzip it.
3. Open `week06-practice.qmd` in RStudio.
4. Keep the `data` folder beside the QMD.
5. Work from top to bottom.
6. For each exercise: **predict → run → inspect → explain**.
7. Edit code and write explanations directly in the QMD.
8. Render frequently.
9. Finish with the **Clean Render Test**.

**Do not copy this webpage into RStudio.** The downloadable QMD is the editable source.

[**Download Week 6 Practice Package (.zip)**](../downloads/practice-workbooks/ANTH418_Week06_Practice.zip)
:::

## How to use this workbook

Week 5 emphasized that visual choices shape what we see. Week 6 goes one step earlier:

> **raw values → preparation decisions → transformed/missing-data representation → analysis → conclusion**

The central claim is:

> **Preparation is analysis.**

Transformations alter mathematical relationships. Missing-data decisions determine which cases count and what an unknown value is allowed to mean.

> **visual audit → diagnose → transform or handle missingness deliberately → compare alternatives → sensitivity analysis → document**

## Learning objectives

By the end, you should be able to diagnose reasons for transformation; compare log, square-root, Box-Cox, and Yeo-Johnson logic; distinguish transformation from z-standardization; explain compositional constraints; distinguish `NA` from zero and explicit from implicit missingness; explain MCAR, MAR, and MNAR; evaluate deletion and imputation; explain multiple imputation; reproduce the core Seshat sensitivity comparison; and document preparation choices.

# Part I — Preparation is analysis

## 1. Cleaning is not neutral bookkeeping

Consider:

1. Replace every unknown outcome with `0`.
2. Apply `log()` to a strongly right-skewed variable.

Both change the data used by later analysis.

### Practice 1

Explain what mathematical structure transformation changes, what evidential claim `NA → 0` makes, and why both decisions should be reported.

# Part II — Why transform a variable?

## 2. Four problems transformations may address

Week 6 identifies:

1. **Skewness correction**
2. **Outlier mitigation**
3. **Variance stabilization**
4. **Linearizing relationships**

A transformation should respond to a diagnosed problem.

## 3. Right skew and tail compression

```{r}
library(tidyverse)

set.seed(418)
raw_x <- rexp(300, rate = 0.08)

transform_demo <- tibble(
  raw = raw_x,
  log = log(raw_x),
  sqrt = sqrt(raw_x)
)

transform_demo |>
  pivot_longer(everything(), names_to = "scale", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ scale, scales = "free") +
  theme_minimal()
```

Compare raw, log, and square-root versions. Which compresses the tail most? Why must interpretation name the transformed scale?

# Part III — Match the function to the data

## 4. Common transformation families

Week 6 introduces log transformations for severe right skew and positive values, roots for counts/milder skew, and reciprocal-type transformations for stronger compression.

```{r}
counts <- c(0, 1, 2, 4, 9, 16, 25)

sqrt(counts)
log(counts)
```

### Practice 2

Why does `log(0)` create a problem? Why is `log(x + 1)` an analytical choice rather than a harmless technical fix?

# Part IV — Box-Cox and Yeo-Johnson

## 5. Search a transformation family

Week 6 presents Box-Cox as a lambda spectrum: approximately no transformation at `λ = 1`, square root at `λ = .5`, log at `λ = 0`, and stronger inverse-style transformations at negative lambda. Box-Cox requires positive data; Yeo-Johnson extends the logic to zero and negative values.

### Practice 3

For values `-2, 0, 3, 7, 20`, explain why ordinary Box-Cox is not directly appropriate and why Yeo-Johnson might be considered.

# Part V — Standardization is not transformation

## 6. Z-scores change units

\[
z = \frac{X-\bar X}{s}
\]

```{r}
measurements <- tibble(
  length_mm = c(40, 45, 50, 55, 60),
  weight_g = c(100, 130, 160, 190, 220)
) |>
  mutate(
    z_length = as.numeric(scale(length_mm)),
    z_weight = as.numeric(scale(weight_g))
  )

measurements
```

```{r}
measurements |>
  summarise(
    mean_z_length = mean(z_length),
    sd_z_length = sd(z_length),
    mean_z_weight = mean(z_weight),
    sd_z_weight = sd(z_weight)
  )
```

Explain why z-scoring changes scale but does not necessarily remove skewness.

# Part VI — Compositional data

## 7. Some data live in constrained spaces

Percentages constrained to sum to a constant can generate artificial correlations. Week 6 introduces log-ratio approaches such as ALR and CLR.

### Practice 4

If an assemblage is 50% A, 30% B, and 20% C, what must happen to another component if A increases while the total remains 100%? Why can closure make ordinary correlation misleading?

# Part VII — Missing values are theoretical objects

## 8. `NA` is not zero

```{r}
x <- c(1, 1, 0, NA, 1, NA, 0)

table(x, useNA = "ifany")
mean(x, na.rm = TRUE)

x_wrong <- replace(x, is.na(x), 0)
mean(x_wrong)
```

### Practice 5

What changed when `NA` became `0`?

> `NA` means ____________________.

> `0` means _____________________.

# Part VIII — Explicit and implicit missingness

## 9. Missing cell versus missing row

```{r}
explicit <- tibble(
  site = c("A", "A", "B", "B"),
  period = c("early", "late", "early", "late"),
  count = c(10, NA, 8, 12)
)

explicit
```

```{r}
implicit <- tibble(
  site = c("A", "B", "B"),
  period = c("early", "early", "late"),
  count = c(10, 8, 12)
)

implicit |>
  complete(site, period)
```

Explain why a cell containing `NA` and an absent row are different data states, and why neither automatically means zero.

# Part IX — Missing-data mechanisms

## 10. MCAR, MAR, MNAR

- **MCAR:** missingness is independent of observed and unobserved values.
- **MAR:** missingness depends on other observed variables.
- **MNAR:** missingness depends on the unobserved value itself or latent causes.

### Practice 6

Classify and justify:

1. A random scanner malfunction deletes 5% of measurements.
2. Older excavation records are more likely to lack coordinates, and record age is known.
3. High-status burials are systematically underreported because access was politically restricted.

Why is mechanism a substantive question rather than something R can determine automatically?

# Part X — Deletion and single imputation have costs

## 11. Listwise deletion changes the sample

```{r}
deletion_demo <- tibble(
  id = 1:8,
  group = c("A","A","A","A","B","B","B","B"),
  x = c(2, 3, NA, 5, 8, 9, NA, 12),
  y = c(4, 5, 6, NA, 10, 11, 12, 13)
)

deletion_demo
deletion_demo |> drop_na(x, y)
```

Which cases disappear? Could deletion alter group composition?

## 12. Mean imputation suppresses variability

```{r}
impute_demo <- c(4, 5, 6, 7, 8, NA, NA)

observed_sd <- sd(impute_demo, na.rm = TRUE)
mean_value <- mean(impute_demo, na.rm = TRUE)
imputed <- replace(impute_demo, is.na(impute_demo), mean_value)

observed_sd
sd(imputed)
```

Why does constant replacement tend to reduce variance?

# Part XI — Multiple imputation preserves uncertainty

## 13. The logic of MICE

Week 6 presents multiple imputation as:

> **incomplete data → several plausible completed datasets → analyze each → pool estimates and standard errors**

Explain why several plausible datasets represent missing-data uncertainty better than one spreadsheet filled with a single mean.

# Part XII — Load the Seshat teaching summary

## 14. Portable data path

```{r}
if (file.exists("data/seshat_missingness_counts.csv")) {
  seshat_file <- "data/seshat_missingness_counts.csv"
} else {
  project_root <- if (file.exists("_quarto.yml")) "." else ".."
  seshat_file <- file.path(
    project_root,
    "assignments",
    "assignment03_seshat_missing",
    "data",
    "seshat_missingness_counts.csv"
  )
}

file.exists(seshat_file)
```

```{r}
seshat <- read_csv(seshat_file, show_col_types = FALSE)

seshat
glimpse(seshat)
names(seshat)
```

::: {.callout-important title="Know the teaching-data unit"}
This CSV is a **three-row aggregate summary**, not 801 individual cases. It summarizes the documented 801-observation Seshat regression frame used for the Week 6 missing-data exercise.
:::

# Part XIII — The documented Seshat states

## 15. Presence, absence, and unknown

The documented regression frame contains:

- **299** known present;
- **12** known absent;
- **490** missing/unknown;
- **801** observations total.

The lecture summarizes this as approximately **37% present, 2% absent, and 61% missing**.

### Practice 7

Inspect the teaching file. Identify its status/category column and count column. Verify that the counts sum to 801.

Because the released teaching file is deliberately aggregated, do not pretend it contains individual histories.

# Part XIV — Preserve unknown as unknown

## 16. Different denominators answer different questions

```{r}
known_present <- 299
known_absent <- 12
missing_unknown <- 490

total <- known_present + known_absent + missing_unknown
known_total <- known_present + known_absent

present_among_known <- known_present / known_total
present_among_all <- known_present / total

present_among_known
present_among_all
```

### Practice 8

Explain why these proportions answer different questions. Which denominator is appropriate depends on the claim being made.

# Part XV — Recode unknown as absence

## 17. Sensitivity to recoding

```{r}
tibble(
  scenario = c(
    "Unknown retained: present among known outcomes",
    "Unknown recoded absent: present among all observations"
  ),
  present_proportion = c(
    present_among_known,
    present_among_all
  )
)
```

```{r}
tibble(
  scenario = c("Known outcomes only", "Unknown → absent"),
  present_proportion = c(present_among_known, present_among_all)
) |>
  ggplot(aes(x = scenario, y = present_proportion)) +
  geom_col() +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = NULL,
    y = "Proportion coded present",
    title = "The descriptive result depends on missing-data treatment"
  ) +
  theme_minimal()
```

Nothing new was observed when `NA` became `0`; the analyst changed its meaning. Explain why comparing these scenarios is a **sensitivity analysis**.

# Part XVI — The Seshat missing-data debate

## 18. Why coding mattered

The Week 6 lecture uses the Seshat debate to show that unknown observations were treated as recorded absences in the disputed analysis. With extensive missingness, this decision strongly shaped the apparent pattern.

> **Recoding unknown as absent imposes a historical claim onto unknown cases.**

The lecture also highlights forward bias: if beliefs can exist before written documentation, coding unrecorded earlier periods as absence can push apparent origins later.

### Practice 9

Explain:

> possible belief → no written record → unknown status

versus

> possible belief → no written record → recoded `0` → apparent absence

Why can the second pathway manufacture a temporal pattern?

# Part XVII — What sensitivity analysis establishes

## 19. Do not overclaim

This exercise shows that the descriptive pattern is highly sensitive to missing-data treatment.

It does **not**, by itself, prove the true value of every missing case, that every missing case should be present, a universal causal sequence, or one uniform missingness mechanism.

### Practice 10

Write two paragraphs:

**A — What the sensitivity analysis demonstrates**

**B — What it does not establish**

# Part XVIII — A practical Week 6 audit

## 20. Rule set before inference

1. **Visual audit** — distributions and missingness.
2. **Diagnose** — skew, variance, nonlinearity, missingness.
3. **Preserve semantics** — never make `NA` mean `0` without evidence.
4. **Choose deliberately** — transformation, deletion, modeling, imputation.
5. **Compare alternatives** — do conclusions change?
6. **Document the pathway** — report preparation decisions.

### Practice 11

Apply this checklist to the Seshat teaching summary. Which steps can be completed with the aggregate file, and which require original case-level data?

# Part XIX — Conceptual debugging

## 21. Diagnose the mistake

Identify the problem:

1. “The data are skewed, so I always log-transform them.”
2. “`log(0)` failed, so I added 1; that has no analytical consequences.”
3. “Z-scoring makes a skewed distribution normal.”
4. “A missing cell is the same as a recorded zero.”
5. “No row means the count must be zero.”
6. “Listwise deletion only reduces sample size.”
7. “Mean imputation is safe because it preserves the mean.”
8. “MICE discovers the true missing values.”
9. “The 490 Seshat unknowns were really absences.”
10. “If two missing-data treatments differ, one is automatically correct.”

# Part XX — Integrated Week 6 challenge

## 22. Build a transparent preparation argument

1. Identify one transformation problem and a plausible transformation.
2. Explain why the transformation fits the diagnosed problem.
3. Distinguish transformation from standardization.
4. Explain `NA ≠ 0`.
5. Distinguish explicit from implicit missingness.
6. Classify one hypothetical MCAR, MAR, and MNAR mechanism.
7. Demonstrate one cost of deletion or single imputation.
8. Reproduce the Seshat sensitivity comparison.
9. Explain how the coding choice changes the descriptive result.
10. State what the aggregate teaching file cannot reveal.
11. Write a short **Data Preparation Decisions** section.
12. Render the document.

# Part XXI — Mastery checklist

- [ ] I treat preparation as part of analysis.
- [ ] I transform only after diagnosing a reason.
- [ ] I understand log, root, Box-Cox, and Yeo-Johnson logic.
- [ ] I distinguish transformation from z-standardization.
- [ ] I recognize compositional constraints.
- [ ] I understand `NA ≠ 0`.
- [ ] I distinguish explicit and implicit missingness.
- [ ] I can explain MCAR, MAR, and MNAR.
- [ ] I understand costs of deletion and single imputation.
- [ ] I understand the logic of multiple imputation.
- [ ] I know the Seshat teaching CSV is a three-row aggregate.
- [ ] I can reproduce the missing-data sensitivity comparison.
- [ ] I distinguish sensitivity from proof.
- [ ] I document preparation choices.

# Final challenge — explain the workflow without code

Explain:

> **inspect → diagnose → preserve meaning → transform/handle missingness → compare alternatives → test sensitivity → document → interpret**

Why is **preparation analysis rather than bookkeeping**?

# Clean Render Test

1. Save the QMD.
2. Choose **Session → Restart R**.
3. Do not recreate objects manually in the Console.
4. Render the complete document.
5. If it fails, identify what the document assumed but never created.

A successful clean render shows that the preparation pathway is computationally self-contained.
