---
title: "ANTH418 Assignment 7: Lithic Raw Materials in Highland Lesotho"
author: "Your Name"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Research question

Does lithic raw-material composition differ among the four published Sehonghong archaeological layers, which layer/material combinations contribute most strongly to the association, and how strong is the association?

## Case study

Gregory, Alex, Peter Mitchell, and Justin Pargeter. 2023. Raw Material Surveys and Their Behavioral Implications in Highland Lesotho. *Journal of Paleolithic Archaeology* 6:18.

```{r}
library(tidyverse)

data_path <- if (file.exists("data/lesotho_raw_material.csv")) {
  "data/lesotho_raw_material.csv"
} else {
  "assignments/assignment07_lesotho_chisquare/data/lesotho_raw_material.csv"
}

dat <- read_csv(
  data_path,
  show_col_types = FALSE
)

count(dat, level)
glimpse(dat)
```

The teaching file contains four source-defined archaeological layers:

- BAS
- RBL-CLBRF
- RF
- BARF

The **unit of analysis** is one lithic observation/artifact record represented in the source flake dataset and retained in this four-layer analytical frame.

Do not interpret 1,060 artifact records as 1,060 independent behavioral events. Artifact frequencies can be affected by reduction, fragmentation, deposition, recovery, and classification.

Do **not** select layers automatically.

## 1. Create the teaching raw-material grouping

Retain these three source categories separately:

- `CrystallineQuartz`
- `Dolerite`
- `FineChert`

Combine:

- `Agate`
- `CoarseChert`
- `Quartzite`
- `Sandstone`

as:

`Other`

```{r}
# Your code
```

`Other` is an **analysis-time teaching grouping**, not an original source category.

Verify that all 1,060 observations remain and that no `material_group` value is missing.

Explain why this transparent collapse is preferable to running Pearson's chi-square test with sparse expected cells.

## 2. Contingency table, proportions, and visualization

Create the **4 × 4 contingency table** of archaeological layer by `material_group`.

Calculate **row proportions**. A row proportion answers:

> **Within this archaeological layer, what proportion of lithic observations belongs to each raw-material group?**

Then create a proportional stacked bar chart, or another appropriate visualization, showing raw-material composition by layer.

```{r}
# Your code
```

Describe the most visible differences before running the formal test. Make the denominator of your proportions explicit.

## 3. Pearson chi-square test

State the null and alternative hypotheses in words.

Run Pearson's chi-square test of independence between:

- archaeological layer; and
- raw-material group.

```{r}
# Your code
```

Before interpreting the p-value, inspect the expected counts.

Report:

- the minimum expected count;
- the number of cells with expected count below 5.

Explain whether the usual chi-square approximation is reasonable for this table.

## 4. Standardized residuals

Inspect the standardized residuals from the chi-square test.

```{r}
# Your code
```

Identify the layer/material combinations that contribute most strongly to the overall association.

As an interpretive guideline, pay particular attention to standardized residuals with absolute magnitude around **2 or greater**.

A positive residual means that a cell occurs more often than expected under independence; a negative residual means that it occurs less often than expected.

Connect the strongest residuals back to the observed row proportions.

## 5. Effect size

Calculate **Cramér's V**.

```{r}
# Your code
```

Use Cramér's V to describe the magnitude of the layer–raw-material association.

Do **not** interpret the chi-square p-value itself as a measure of effect size.

## 6. Interpretation and sparse-table reasoning

Explain the association archaeologically.

When would Fisher's exact test or another sparse-table strategy be more appropriate?

Why is Fisher's exact test **not necessary here**, given the expected counts in this teaching table?

Compare your statistical result with Gregory et al.'s archaeological interpretation. Explain what the layer–raw-material association supports descriptively and statistically, and what additional evidence would be needed to argue for procurement strategy, mobility, climatic causation, or social interaction.

## Final interpretation

Answer the following questions in complete sentences:

1. What is the unit of analysis?
2. How does raw-material composition vary among BAS, RBL-CLBRF, RF, and BARF?
3. What does the chi-square test tell you about association between archaeological layer and raw-material composition?
4. Are the expected counts adequate for the ordinary chi-square approximation?
5. Which layer/material combinations contribute most strongly, and in what direction?
6. What does Cramér's V add beyond the p-value?
7. Why is Fisher's exact test not required for this validated teaching table, and when would it become useful?
8. What behavioral, procurement, mobility, climatic, or social claim does this association **not** establish by itself?
