---
title: "ANTH418 Assignment 5: Teotihuacan Leporid Management"
author: "Your Name"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Research question

How large is the δ13C difference between leporids from **Oztoyahualco** and the **Moon Pyramid**, how uncertain is it, and how unusual would that difference be under a no-difference null model?

This assignment follows the Week 8 workflow:

> **Estimate → quantify uncertainty → test against a null model → interpret archaeologically**

## Case study

Somerville, Andrew D., Nawa Sugiyama, Linda R. Manzanilla, and Margaret J. Schoeninger. 2016. Animal Management at the Ancient Metropolis of Teotihuacan, Mexico: Stable Isotope Analysis of Leporid (Cottontail and Jackrabbit) Bone Mineral. *PLOS ONE* 11(8): e0159982.

This two-context comparison is a teaching exercise using published observations; it is not a reproduction of the full statistical analysis in Somerville et al. (2016).

Before beginning, read `README.md`, `codebook.md`, and `provenance.md`.

```{r}
library(tidyverse)
library(infer)

data_path <- if (file.exists("data/teotihuacan_leporids.csv")) {
  "data/teotihuacan_leporids.csv"
} else {
  "assignments/assignment05_teotihuacan_inference/data/teotihuacan_leporids.csv"
}

dat <- read_csv(
  data_path,
  show_col_types = FALSE
) |>
  filter(valid)

d <- dat |>
  filter(
    context %in% c(
      "Oztoyahualco",
      "Moon Pyramid"
    )
  )

d |> count(context)

glimpse(d)
```

For this assignment, `valid` refers specifically to the **δ13C exclusion rule** documented in the source supporting information.

The effect is always defined as:

**Oztoyahualco − Moon Pyramid**

A positive difference therefore means that mean δ13C is higher (less negative) at Oztoyahualco.

---

## 1. Inspect and prepare

Identify:

- the observational unit;
- the grouping variable;
- the numerical response variable;
- the sample size in each context.

In one or two sentences, state what archaeological comparison these observations can reasonably inform.

Remember that these archaeological specimens are not automatically a random probability sample of all Teotihuacan leporids.

## 2. Describe and visualize the observed groups

Calculate group `n`, mean, and SD for δ13C.

Create one clear visualization comparing the two observed distributions. Show enough of the underlying observations to assess overlap and spread.

```{r}
# Your code
```

Briefly describe:

- the direction of the observed difference;
- overlap and spread;
- any notable features of the distributions.

Do not begin with statistical significance. Begin with the observed data.

## 3. Estimate the observed difference

Calculate the mean δ13C difference as:

**Oztoyahualco − Moon Pyramid**

```{r}
# Your code
```

Report the estimate in ‰.

Explain:

1. what its sign means;
2. how large the observed contrast is; and
3. what this point estimate does **not** tell you about uncertainty.

## 4. Bootstrap confidence interval

Set the random seed:

```{r}
set.seed(418)
```

Using `infer`, generate **10,000 bootstrap resamples** and calculate the difference in means in the same direction:

**Oztoyahualco − Moon Pyramid**

Construct a **95% percentile bootstrap confidence interval**.

```{r}
# Your code using infer
```

Visualize the bootstrap distribution and mark the confidence interval.

Report:

- the observed mean difference;
- lower confidence bound;
- upper confidence bound.

Interpret the interval in terms of **magnitude and uncertainty**.

Do not say that there is a 95% probability that the fixed population difference lies inside this particular frequentist interval.

Also explain why the bootstrap distribution is a distribution of **estimates**, not a distribution of individual leporid δ13C values.

## 5. Specify the null model

For this comparison, use:

$$
H_0:
\mu_{\text{Oztoyahualco}}
-
\mu_{\text{Moon Pyramid}}
=0
$$

and the two-sided alternative:

$$
H_A:
\mu_{\text{Oztoyahualco}}
-
\mu_{\text{Moon Pyramid}}
\neq0.
$$

In words, under the null model, context carries no systematic information about δ13C for this comparison.

Explain why this statistical null is **not** equivalent to saying that no leporid management occurred at Teotihuacan.

## 6. Permutation null distribution

Reset the seed:

```{r}
set.seed(418)
```

Under the no-association null model, use `infer` to generate **10,000 permutations** of the context labels.

Calculate the difference in means for every permutation.

```{r}
# Your code using infer
```

Visualize the **null distribution** and mark the observed difference.

Calculate a **two-sided permutation p-value** using permuted statistics at least as extreme as the observed statistic.

## 7. Interpret the p-value

Report the p-value as a continuous value.

Explain it as a **tail probability conditional on the null model**.

Your interpretation must make clear that the p-value is **not**:

- the probability that the null hypothesis is true;
- the probability that leporid management occurred;
- the probability that chance caused the result;
- the magnitude of the isotope difference; or
- a measure of archaeological importance.

## 8. Integrate estimation and testing

Interpret together:

1. the observed difference in means;
2. the 95% bootstrap confidence interval;
3. the two-sided permutation p-value.

Answer:

- How large is the observed contrast?
- How uncertain is the estimate?
- What range of differences is compatible with the data under the bootstrap procedure?
- How unusual is the observed difference under the permutation null model?
- Do the estimate, interval, and null-model comparison tell a coherent story?

The confidence interval and p-value answer **different questions**. Do not let the p-value replace the effect estimate or its uncertainty.

## 9. Return to archaeology

Relate the statistical evidence to Somerville et al. (2016).

Discuss:

- what higher δ13C may imply about C4/CAM consumption;
- why access to cultivated foods may be consistent with human provisioning;
- why provisioning is not automatically equivalent to management or breeding;
- how archaeological context contributes to the interpretation; and
- at least one relevant limitation or alternative explanation.

Possible limitations include sampling and preservation, contextual or chronological structure, independence of observations, environmental or dietary baselines, and the possibility that hunting and management occurred together.

Distinguish clearly between:

> **what the statistical analysis establishes about the isotope contrast**

and

> **what the broader archaeological evidence suggests about human–leporid relationships.**

## Final interpretation

Answer the research question in complete sentences.

Your conclusion should integrate:

- effect magnitude;
- uncertainty;
- the null-model comparison;
- archaeological significance;
- limitations.

Do not end with only "reject H0" or "statistically significant."

> **Estimate first. Quantify uncertainty. Test a clearly defined null model when useful. Then return to archaeology.**

## Scope

You are **not** required to reproduce all statistical procedures used by Somerville et al. (2016), or to perform ANOVA, Mann–Whitney tests, t-tests, Bayesian analysis, or causal modeling.

Use only the Week 8 methods required here.