---
title: "ANTH418 Assignment 2: Visualizing Kinship Diversity with Kinbank"
author: "Your Name"
format:
  html:
    toc: true
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-note title="Before you begin"}
Read the **Week 5 Study Guide** and the Assignment 2 instructions before starting. This assignment is limited to **exploratory data analysis and visualization**. The goal is to learn what the data show, identify questions worth investigating, and recognize what the visualizations alone cannot establish.
:::

## Case Study

Kinbank is a global database of kinship terminology designed for systematic cross-cultural comparison (Passmore et al., 2023). The ANTH418 teaching dataset is a frozen later snapshot containing **1,287 languages represented by at least one Kinbank form record**. Each row represents one language.

The variable `kin_term_count` is the number of Kinbank **form records** linked to a language in this teaching snapshot. It is not automatically the number of unique kin concepts and should not be treated as a direct measure of kinship-system complexity.

## Research question

**What can exploratory visualization reveal about the structure of the Kinbank teaching sample and variation in `kin_term_count`, and what dependencies or limitations must be considered before moving from description to inference?**

## 1. Inspect the data

Load the dataset and inspect its structure.

```{r}
library(tidyverse)

dat <- read_csv("assignments/assignment02_kinbank_visualization/data/kinbank_teaching.csv", show_col_types = FALSE)

glimpse(dat)
```

Confirm the number of observations and variables, identify categorical and numerical variables, and check missingness in the variables you will use.

```{r}
# Your code
```

**Write briefly:**

- What does one row represent?
- Which variables are categorical and which are numerical?
- What exactly does `kin_term_count` count?
- Which relevant variables contain missing values?

## 2. Visualize the structure of the evidence

Before interpreting kinship diversity, examine **who is represented in the sample**.

Create one clear **bar chart** showing the number of languages by either `family` or `macroarea`. If you use `family`, choose a manageable, clearly justified set of well-represented families rather than producing an unreadable plot.

```{r}
# Your code
```

**Interpret:** Which groups are most and least represented? Is representation balanced or uneven? Why does this matter for later comparisons?

## 3. Examine the distribution of `kin_term_count`

Create a histogram of `kin_term_count`.

```{r}
# Your code
```

During exploration, try at least one alternative reasonable bin width before choosing the version you will emphasize.

```{r}
# Your exploratory code
```

**Interpret:** Describe relevant features of the distribution—center, spread, skewness, modes, gaps, and unusual observations. Explain briefly whether changing bin width altered what was visually prominent.

## 4. Make one purposeful comparison across groups

Choose **one** comparison of `kin_term_count` across either `macroarea` or a manageable set of `family` groups.

Use a graphic that matches your question. Suitable choices include a boxplot (with raw observations where useful) or faceted histograms.

```{r}
# Your code
```

**Explain:**

1. Why is this graphic appropriate for the variables and comparison?
2. What similarities and differences are visible **in this sample**?
3. Do unequal group sizes or graphical choices affect the comparison?

If you use facets, fixed scales are generally preferable when direct between-panel comparison is the goal.

## 5. Consider non-independence

In a short paragraph, explain how **phylogenetic autocorrelation**, **spatial autocorrelation**, or both could affect interpretation of the patterns you visualized.

Related languages may resemble one another because of shared ancestry and inherited terminology. Geographically close languages may resemble one another because of contact, borrowing, regional history, or shared environments.

You are **not** required to calculate autocorrelation, construct a phylogenetic tree, or fit a phylogenetic model. Visualization can make dependence visible; it does not correct for dependence.

## 6. Final EDA interpretation

Conclude by answering all three questions:

1. **What do your visualizations show about the observed Kinbank teaching sample?**
2. **What is one question or apparent pattern worth investigating further?**
3. **What is one conclusion these visualizations alone do not justify?**

Use descriptive language such as *in this sample*, *the distribution shows*, *the graph suggests*, and *worth investigating*. Do not claim that the graphs prove a universal cross-cultural pattern, causal relationship, or evolutionary process.

::: {.callout-important title="Statistical scope"}
No hypothesis tests, p-values, confidence intervals, causal analyses, phylogenetic comparative models, or spatial autocorrelation tests are required.
:::

## Submission

Submit both:

1. Completed Quarto source file (`.qmd`)
2. Rendered HTML file (`.html`)

Before submitting, restart R and render the document from a clean session.

## References

Passmore, S., Barth, W., Greenhill, S. J., Quinn, K., Sheard, C., Argyriou, P., et al. (2023). Kinbank: A global database of kinship terminology. *PLOS ONE, 18*(5), e0283218. https://doi.org/10.1371/journal.pone.0283218

Passmore, S., & Jordan, F. M. (2020). No universals in the cultural evolution of kinship terminology. *Evolutionary Human Sciences, 2*, e42. https://doi.org/10.1017/ehs.2020.41
